The set of real numbers, denoted \( \mathbb{R} \), includes all rational and irrational numbers and forms a continuous, uncountable set. While theoretically infinite, practical computations in digital systems can only represent a finite subset of \( \mathbb{R} \). Finite representations of real numbers are constrained by the limitations of hardware and software, leading to approximations in many cases. 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Definition and Characteristics}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Finite real numbers are those real numbers that can be precisely represented within a given computational framework or mathematical system. For instance, within the context of floating-point arithmetic, a finite real number must satisfy:

\[
x \in \mathbb{R} \quad \text{and} \quad \text{finite precision representation exists}.
\]

Unlike integers, which are discrete, real numbers encompass rational values (e.g., \( \frac{1}{2} \), \( 0.75 \)) and irrational values (e.g., \( \sqrt{2} \), \( \pi \)). Finite real numbers are typically confined to the subset of \( \mathbb{R} \) that can be approximated by rational representations due to hardware precision constraints.

In computation, real numbers are often represented using floating-point or fixed-point formats:

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Floating Point Representation of Real Numbers}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The IEEE 754 standard defines floating-point formats, which provide an efficient and scalable way to approximate real numbers using a finite number of bits. These formats are widely used in scientific computations, graphics, and machine learning due to their ability to balance precision, range, and efficiency.

%********************************************
\subsection{General Floating-Point Format}
%********************************************
A floating-point number under the IEEE 754 standard \cite{ieee2008} is expressed as:

\[
x = (-1)^s \cdot m \cdot 2^e,
\]

where:
\begin{itemize}[noitemsep]
    \item \( s \): The sign bit, which determines whether the number is positive (\( s = 0 \)) or negative (\( s = 1 \)).
    \item \( m \): The significand (or mantissa), which represents the precision of the number. It is normalized to the form \( 1.f \), where \( f \) is the fractional part in binary.
    \item \( e \): The exponent, which determines the scale or range of the number. The exponent is stored in a biased format to allow both positive and negative exponents.
\end{itemize}

For example, the decimal number \( 6.25 \) is represented as:

\[
6.25 = 1.5625 \cdot 2^2 \quad \text{(binary: \( 1.101 \cdot 2^2 \))}.
\]

In a floating-point format, \( s = 0 \), \( m = 1.101 \), and \( e = 2 \) (adjusted with the exponent bias).

%***************************************************
\subsection{Rounding in Floating-Point Arithmetic}
%****************************************************
Finite precision requires approximating real numbers (\( \mathbb{R} \)) with a subset of representable values. Floating-point numbers have a finite number of bits, limiting the precision with which real numbers can be represented. When a computation produces a result that cannot be represented exactly (e.g., \( 1/3 \) in binary), rounding is applied to map the result to the nearest representable value.

Without consistent rounding rules, numerical computations could yield unpredictable results, especially in iterative algorithms or when combining multiple operations. The IEEE 754 standard specifies several rounding modes to handle this approximation, each designed to suit specific computational needs and numerical properties.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{IEEE 754 Rounding Modes.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The IEEE 754 standard defines the following five rounding modes:

\begin{enumerate}
    \item Round to Nearest, Ties to Even (Default):
    \begin{itemize}
        \item The value is rounded to the nearest representable number.
        \item If the result is exactly halfway between two representable values, it is rounded to the nearest even value.
        \item Use Case: This is the default rounding mode and minimizes overall rounding error in repeated computations.
    \end{itemize}
    
    \item Round to Nearest, Ties Away from Zero:
    \begin{itemize}
        \item The value is rounded to the nearest representable number.
        \item If the result is exactly halfway between two representable values, it is rounded away from zero.
        \item Use Case: Useful in financial applications where avoiding bias toward zero is important.
    \end{itemize}
    
    \item Round Toward Zero (Truncation):
    \begin{itemize}
        \item The fractional part is discarded, effectively truncating the value toward zero.
        \item Use Case: Often used in integer arithmetic conversions to avoid exceeding bounds.
    \end{itemize}
    
    \item Round Toward Positive Infinity (\(+\infty\)):
    \begin{itemize}
        \item The value is rounded up to the nearest representable number.
        \item Use Case: Useful in applications requiring upper bounds, such as interval arithmetic or error estimation.
    \end{itemize}
    
    \item Round Toward Negative Infinity (\(-\infty\)):
    \begin{itemize}
        \item The value is rounded down to the nearest representable number.
        \item Use Case: Useful in applications requiring lower bounds or ensuring conservative estimates.
    \end{itemize}
\end{enumerate}

%**************************************
\paragraph{Examples of Rounding Modes.}
%**************************************
Consider rounding the number \( x = 2.625 \) in binary (\( 10.101_2 \)) to 3 bits of precision in the fractional part.

\begin{enumerate}
    \item Round to Nearest, Ties to Even:
     \( x \) rounds to \( 10.10_2 \) (2.5 in decimal) because it is closer to \( 10.10_2 \) than \( 10.11_2 \).
    \item Round to Nearest, Ties Away from Zero:
     \( x \) rounds to \( 10.11_2 \) (2.75 in decimal).
    \item Round Toward Zero:
     \( x \) rounds to \( 10.10_2 \) (2.5 in decimal) by truncating the fractional part.
    \item Round Toward Positive Infinity:
     \( x \) rounds to \( 10.11_2 \) (2.75 in decimal).
    \item Round Toward Negative Infinity:
     \( x \) rounds to \( 10.10_2 \) (2.5 in decimal).
\end{enumerate}

%*************************************
\paragraph{Summary of Rounding Modes.}
%*************************************
Table \ref{tab:rounding_modes} shows the IEEE-754 rounding modes.
\begin{table}[h!]
\centering
\renewcommand{\arraystretch}{1.3} % Adjust row height for readability
\setlength{\tabcolsep}{6pt} % Adjust column spacing
\caption{IEEE 754 Rounding Modes}
\label{tab:rounding_modes}
\begin{tabular}{|p{4cm}|p{8cm}|}
\hline
\textbf{Rounding Mode}                 & \textbf{Description and Use Case}                                                                                     \\ \hline
Round to Nearest, Ties to Even         & Rounds to the nearest representable number; ties go to the nearest even number. Default mode for minimizing bias.     \\ \hline
Round to Nearest, Ties Away from Zero  & Rounds to the nearest representable number; ties go away from zero. Used in financial and accounting applications.     \\ \hline
Round Toward Zero                      & Discards the fractional part, effectively truncating the value. Useful in integer conversions.                         \\ \hline
Round Toward Positive Infinity (\(+\infty\)) & Rounds up to the nearest representable number. Useful for bounding errors or interval arithmetic.                     \\ \hline
Round Toward Negative Infinity (\(-\infty\)) & Rounds down to the nearest representable number. Useful for conservative lower bounds.                                 \\ \hline
\end{tabular}
\end{table}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The Role of Denormalized Numbers in Floating-Point Arithmetic}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Floating-point systems, such as those defined by the IEEE 754 standard, provide an approximation of the real number line (\( \mathbb{R} \)) within a finite precision. This approximation forms a discrete metric space, where the representable numbers are distributed non-uniformly. Denormalized numbers (also called subnormal numbers) are introduced to address the issue of spacing near zero, improving the system's ability to approximate \( \mathbb{R} \).

%******************************************************
\paragraph{Discrete Spacing in Standard Floating-Point.}
%*******************************************************

In normalized floating-point representation, the spacing between consecutive representable numbers is determined by the precision (significand bits) and the magnitude of the exponent. For a normalized floating-point number:

\[
x = (-1)^s \cdot (1.f) \cdot 2^e,
\]

The discrete spacing between successive numbers doubles as the magnitude of \( x \) increases. For example, near zero (for small exponents), the numbers are closely packed, but as \( e \) grows, the spacing becomes increasingly larger. This doubling of spacing is a direct result of the binary scaling factor \( 2^e \) applied in the IEEE 754 standard. This spacing leaves a gap near zero, creating a discontinuity in the metric space.

\vspace{\baselineskip}\par
\noindent Consider a simplified 4-bit floating-point system with the following structure:
\begin{itemize}
    \item Sign bit (\( s \)): 1 bit.
    \item Exponent (\( e \)): 2 bits, biased by 1 (\( \text{bias} = 1 \)).
    \item Fraction (mantissa) (\( f \)): 1 bit.
\end{itemize}

The value of a normalized number is given by:
\[
x = (-1)^s \cdot (1.f) \cdot 2^{e - \text{bias}}.
\]

In this system:
\begin{itemize}
    \item The exponent range is \( e \in [0, 3] \), corresponding to actual exponents \( e - 1 \in [-1, 2] \).
    \item The significand range is \( 1.0 \) or \( 1.5 \) (depending on \( f \)).
\end{itemize}

We now show how the spacing changes. For each exponent value, the spacing doubles as \( e \) increases:

\begin{itemize}
    \item Exponent \( e = 0 \) (\( 2^{-1} \)):
    \[
    x \in \{1.0 \cdot 2^{-1} = 0.5, \; 1.5 \cdot 2^{-1} = 0.75\}.
    \]
    Spacing: \( 0.75 - 0.5 = 0.25 \).

    \item Exponent \( e = 1 \) (\( 2^{0} \)):
    \[
    x \in \{1.0 \cdot 2^0 = 1.0, \; 1.5 \cdot 2^0 = 1.5\}.
    \]
    Spacing: \( 1.5 - 1.0 = 0.5 \).

    \item Exponent \( e = 2 \) (\( 2^{1} \)):
    \[
    x \in \{1.0 \cdot 2^1 = 2.0, \; 1.5 \cdot 2^1 = 3.0\}.
    \]
    Spacing: \( 3.0 - 2.0 = 1.0 \).
\end{itemize}

As the exponent increases, the spacing between consecutive representable numbers doubles:
\begin{itemize}
    \item For \( e = 0 \), spacing is \( 0.25 \).
    \item For \( e = 1 \), spacing is \( 0.5 \).
    \item For \( e = 2 \), spacing is \( 1.0 \).
\end{itemize}

This behavior reflects the structure of normalized floating-point numbers, where the range of the significand remains fixed, but the scaling factor \( 2^e \) determines the magnitude and spacing.

The doubling of spacing as the exponent increases reduces the precision of floating-point representation for larger magnitudes. While this allows for a wide dynamic range, it introduces rounding errors in computations involving large and small numbers.

%*********************************
\paragraph{Limitations Near Zero.}
%*********************************
When the exponent \( e \) reaches its minimum value (e.g., \( e_{\text{min}} = -126 \) for single precision), the smallest normalized floating-point number is:

\[
x_{\text{min}} = 1.0 \cdot 2^{e_{\text{min}}}.
\]

Numbers smaller than \( x_{\text{min}} \) cannot be represented in the normalized format, leaving a gap between \( 0 \) and \( x_{\text{min}} \). This discontinuity can cause issues in numerical computations, such as abrupt underflow to zero, which leads to a loss of precision in operations involving small values.

%*********************************************************
\paragraph{Denormalized Numbers Improve the Metric Space.}
%*********************************************************
Denormalized numbers are introduced to fill the gap near zero by allowing a leading significand bit of \( 0 \), instead of the implicit \( 1 \) required for normalized numbers. A denormalized number is represented as:

\[
x = (-1)^s \cdot (0.f) \cdot 2^{e_{\text{min}}}.
\]

This representation permits a gradual decrease in the magnitude of \( x \) as the fractional part \( f \) decreases. For single precision, the range of denormalized numbers is:

\[
x \in \left[ 2^{-149}, 2^{-126} \right].
\]

By introducing smaller spacings near zero, denormalized numbers extend the metric space of floating-point numbers, enabling a more continuous approximation of \( \mathbb{R} \).

Consider a simplified 4-bit floating-point system with:
\begin{itemize}
    \item Sign bit (\( s \)): 1 bit.
    \item Exponent (\( e \)): 2 bits, biased by 1 (\( \text{bias} = 1 \)).
    \item Fraction (mantissa) (\( f \)): 1 bit.
\end{itemize}

In this system, denormalized numbers allow the leading bit of the significand to be \( 0 \), enabling the representation of smaller values close to zero. The value of a denormalized number is:

\[
x = (-1)^s \cdot (0.f) \cdot 2^{e_{\text{min}}}.
\]

Representation of Denormalized Numbers
For \( e = 0 \) (the smallest exponent, used for denormalized numbers), the representable values are:

\begin{itemize}
    \item Fraction \( f = 0.0 \):
    \(
    x = 0.0 \cdot 2^{-1} = 0.
    \)

    \item Fraction \( f = 0.25 \):
    \(
    x = 0.25 \cdot 2^{-1} = 0.125.
    \)

    \item Fraction \( f = 0.5 \):
    \(
    x = 0.5 \cdot 2^{-1} = 0.25.
    \)

    \item Fraction \( f = 0.75 \):
    \(
    x = 0.75 \cdot 2^{-1} = 0.375.
    \)
\end{itemize}

\vspace{\baselineskip}\par
\noindent 
The spacing between consecutive denormalized numbers remains constant:
\begin{itemize}
    \item Between \( 0.0 \) and \( 0.125 \): Spacing is \( 0.125 \).
    \item Between \( 0.125 \) and \( 0.25 \): Spacing is \( 0.125 \).
    \item Between \( 0.25 \) and \( 0.375 \): Spacing is \( 0.125 \).
\end{itemize}

Table \ref{tab:denormalized_spacing} shows the denormalized spacing for IEEE formats. The inclusion of denormalized numbers affects the spacing within the floating-point metric space:
\begin{itemize}
    \item Better Approximation Near Zero: Denormalized numbers significantly improve the precision of calculations involving very small values, reducing the abrupt transition to zero caused by underflow.
    \item Smooth Gradation: The gradual decrease in representable values leads to smoother behavior in numerical algorithms, particularly in iterative methods that rely on small updates.
    \item Support for Gradual Underflow: Instead of truncating to zero, operations involving small values can produce denormalized results, preserving information that would otherwise be lost.
\end{itemize}

\begin{table}[h!]
\centering
\renewcommand{\arraystretch}{1.3} % Adjust row height for readability
\setlength{\tabcolsep}{6pt} % Adjust column spacing for better layout
\caption{Denormalized Spacing in IEEE 754 Floating-Point Formats}
\label{tab:denormalized_spacing}
\begin{tabular}{|p{3.5cm}|p{4cm}|p{4cm}|p{1.5cm}|}
\hline
\textbf{Format}          & \textbf{Minimum Denormalized Value} & \textbf{Next Larger Denormalized Value} & \textbf{Spacing} \\ \hline
Half-Precision (f16)     & \( 2^{-24} \approx 5.96 \times 10^{-8} \) & \( 2^{-23} \approx 1.19 \times 10^{-7} \) & \( 2^{-24} \) \\ \hline
Single-Precision (f32)   & \( 2^{-149} \approx 1.4 \times 10^{-45} \) & \( 2^{-148} \approx 2.8 \times 10^{-45} \) & \( 2^{-149} \) \\ \hline
Double-Precision (f64)   & \( 2^{-1074} \approx 4.9 \times 10^{-324} \) & \( 2^{-1073} \approx 9.9 \times 10^{-324} \) & \( 2^{-1074} \) \\ \hline
Quadruple-Precision (f128) & \( 2^{-16494} \approx 6.5 \times 10^{-4966} \) & \( 2^{-16493} \approx 1.3 \times 10^{-4965} \) & \( 2^{-16494} \) \\ \hline
\end{tabular}
\end{table}


Consider a single-precision system (\( e_{\text{min}} = -126 \)):
\begin{itemize}[noitemsep]
    \item Without Denormalized Numbers: The smallest representable positive number is \( 2^{-126} \), with a large gap to zero.
    \item With Denormalized Numbers: Values like \( 2^{-127}, 2^{-128}, \dots, 2^{-149} \) are introduced, reducing the spacing and providing a closer approximation to \( \mathbb{R} \).
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{IEEE Supported Formats}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
IEEE-754 and later editions support many different formats. We look at f16 in detail due to the ease of displaying bit widths. However, all formats work similarly. \ref{tab:ieee_precisions} shows the IEEE format definitions. Note that Decimal representations use Densely Packed Decimal (DPD) representations which require $n \times \frac{10}{3}$ bits for n-digits rather than standard BCD which would require $n \times 4$ bits per digit.

\begin{table}[h!]
\centering
\renewcommand{\arraystretch}{1.5} % Adjust row height for readability
\setlength{\tabcolsep}{5pt} % Adjust column separation for spacing
\caption{Summary of IEEE 754 Supported Floating-Point Precisions}
\label{tab:ieee_precisions}
\begin{tabular}{|p{3cm}|>{\raggedleft\arraybackslash}p{1cm}|>{\raggedleft\arraybackslash}p{2cm}|>{\raggedleft\arraybackslash}p{2cm}|p{5cm}|}
\hline
\textbf{Format}            & \textbf{Total Bits} & \textbf{Exponent Bits} & \textbf{Significand Bits} & \textbf{Applications}                           \\ \hline
Half-Precision (f16)       & 16            & 5                      & 10                        & Graphics, machine learning                      \\ \hline
Single-Precision (f32)     & 32            & 8                      & 23                        & Scientific computing, gaming                    \\ \hline
Double-Precision (f64)     & 64            & 11                     & 52                        & High-precision engineering, financial work      \\ \hline
Quadruple-Precision (f128) & 128           & 15                     & 112                       & Cryptography, scientific research               \\ \hline
Decimal32                  & 32            & 8                      & 7 digits                  & Financial applications                          \\ \hline
Decimal64                  & 64            & 11                    & 16 digits                 & Financial applications                          \\ \hline
Decimal128                 & 128           & 15                    & 34 digits                 & Precision-critical financial applications       \\ \hline
\end{tabular}
\end{table}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Half-Precision Floating-Point (f16)}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The IEEE 754 half-precision (f16) format is a compact 16-bit representation, commonly used in graphics, machine learning, and resource-constrained environments. It is defined as follows:

\begin{itemize}
    \item Bit Layout: A 16-bit number is divided into three components:
    \begin{center}
    \texttt{Sign (1 bit) | Exponent (5 bits) | Fraction (10 bits)}
    \end{center}
    \item Sign Bit (\( s \)): The first bit determines the sign of the number (0 for positive, 1 for negative).
    \item Exponent (\( e \)): The next 5 bits represent the exponent in a biased format. The bias for f16 is \( 2^{k-1} - 1 = 15 \) (where \( k = 5 \)).
    \item Fraction (\( f \)): The final 10 bits store the fractional part of the significand. The leading bit of the significand is implicitly \( 1 \) for normalized numbers.
\end{itemize}

\paragraph{Value Calculation:}
The value of an f16 floating-point number is calculated as:
\[
x = (-1)^s \cdot (1.f) \cdot 2^{e - 15},
\]
where \( s \) is the sign bit, \( f \) is the fraction, and \( e \) is the biased exponent.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Special Cases in f16:}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{itemize}
    \item Normalized Numbers: For \( 1 \leq e \leq 30 \), the number is normalized, and the leading bit of the significand is \( 1 \). The value is:
    \[
    x = (-1)^s \cdot (1.f) \cdot 2^{e - 15}.
    \]
    \item Denormalized Numbers: For \( e = 0 \), the number is denormalized (used to represent values close to zero). The leading bit of the significand is \( 0 \), and the value is:
    \[
    x = (-1)^s \cdot (0.f) \cdot 2^{-14}.
    \]
    \item Zero: When \( e = 0 \) and \( f = 0 \), the value is \( x = 0 \) (with a sign determined by \( s \)).
    \item Infinity: When \( e = 31 \) and \( f = 0 \), the value is \( x = \pm \infty \) (depending on the sign bit \( s \)).
    \item NaN (Not a Number): When \( e = 31 \) and \( f \neq 0 \), the value is NaN, used for undefined operations (e.g., \( 0/0 \)).
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Example: Representing 6.25 in f16}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

To represent \( 6.25 \) in f16:
\begin{enumerate}
    \item Convert \( 6.25 \) to binary: \( 6.25 = 1.101 \cdot 2^2 \).
    \item Determine the components:
    \begin{itemize}[noitemsep]
        \item \( s = 0 \) (positive number).
        \item \( e = 2 + 15 = 17 \) (add the bias of 15).
        \item \( f = 1010000000_2 \) (fractional part padded to 10 bits).
    \end{itemize}
    \item Write the bit layout:
    \[
    \texttt{0 | 10001 | 1010000000}.
    \]
    \item Verify:
    \[
    x = (-1)^0 \cdot (1.101) \cdot 2^{17 - 15} = 6.25.
    \]
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Applications and Trade-Offs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The f16 format is widely used in applications where memory and computational efficiency are important:
\begin{itemize}[noitemsep]
    \item Graphics: In rendering pipelines, f16 provides sufficient precision for colors, lighting, and textures.
    \item Machine Learning: Many AI accelerators and frameworks (e.g., TensorFlow, PyTorch) use half-precision floating-point formats like IEEE f16 or bfloat16 for faster training and inference.
    \item Embedded Systems: The compact size of f16 makes it ideal for low-power devices with limited memory.
\end{itemize}

However, the limited precision of f16 can lead to significant rounding errors in calculations, particularly for values requiring high precision or a large dynamic range \cite{goldberg1991, ieee2008}.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Fixed-Point Representation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Fixed-point representation is a method for storing and processing numbers by dividing a fixed number of bits into integer and fractional parts. This approach enables efficient computations, particularly in resource-constrained environments such as embedded systems, digital signal processing (DSP), and real-time control systems \cite{lathi2005digital}.

Compared to floating-point representation, fixed-point arithmetic offers simplicity, lower computational overhead, and predictable precision. However, it comes with limitations in dynamic range and precision.

%********************************************
\subsection{Structure of Fixed-Point Numbers}
%*********************************************
In fixed-point representation, a number is divided into two parts:
\begin{itemize}[noitemsep]
    \item Integer Part: Specifies the whole number portion.
    \item Fractional Part: Specifies the fractional portion using a fixed scaling factor.
\end{itemize}

The scaling factor \( 2^q \) determines how the binary digits in the fractional part are interpreted. For an \( n \)-bit representation with \( p \) bits for the integer part and \( q \) bits for the fractional part:
\[
\text{Value} = (-1)^s \cdot \left( \text{Integer Part} + \frac{\text{Fractional Part}}{2^q} \right),
\]
where \( s \) is the sign bit for signed numbers.

For example, in an 8-bit fixed-point format with 4 integer bits and 4 fractional bits (\( q = 4 \)):
\begin{itemize}[noitemsep]
    \item \( 6.25 \) is represented as \( 0110.0100_2 \), with \( 0110_2 \) as the integer part and \( 0100_2 \) as the fractional part.
    \item The scaling factor is \( 2^4 = 16 \), so \( 6.25 \) is stored as \( 6 + \frac{4}{16} = 6.25 \).
\end{itemize}


%*********************************************
\subsection{Range of Signed Fixed-Point Representation}
%*********************************************
Signed fixed-point numbers include a sign bit to represent negative values. The sign bit occupies the most significant bit (MSB), and the remaining bits are divided into integer and fractional parts. Negative numbers are typically represented using two's complement because it allows the same hardware to be used as with 2's complement integers.

For example, in an 8-bit signed fixed-point system with 4 fractional bits:
\begin{itemize}[noitemsep]
    \item \( 6.25 \): \( 0110.0100_2 \) (same as the unsigned case).
    \item \( -6.25 \): Two's complement of \( 0110.0100_2 \) is \( 1001.1100_2 \).
\end{itemize}

This system supports a range of values:
\[
\text{Range} = \left[ -2^{p-1}, 2^{p-1} - \frac{1}{2^q} \right].
\]
For the 8-bit system with \( p = 4 \) and \( q = 4 \), the range is:
\[
[-8.000, +7.9375].
\]

%************************************************
\subsection{Properties of Fixed-Point Arithmetic}
%************************************************

Fixed-point arithmetic has distinct properties that make it well-suited for certain applications. These properties arise from its representation, where numbers are stored as integers with an implied binary point. 

%******************************************
\paragraph{Fixed Precision and Scaling.}
%******************************************

Fixed-point arithmetic divides a number into two parts: an integer part \( X \in \mathbb{Z} \) and a fractional part, allowing it to approximate real numbers \( x \in \mathbb{R} \). A predetermined scaling factor \( 2^q \), where \( q \) is the number of fractional bits, determines the granularity of the representation. Each fixed-point number is expressed as:
\[
x = \frac{X}{2^q}, \quad \text{where } X \in \mathbb{Z} \text{ and } x \in \mathbb{R}.
\]

This representation allows precise mapping between integer values and their scaled equivalents in the real number domain:
\begin{itemize}
    \item The integer part determines the range of representable values.
    \item The fractional part determines the smallest difference (\( 2^{-q} \)) between two consecutive representable values, ensuring a fixed level of precision.

\end{itemize}

\begin{itemize}
    \item Precision remains constant for all values within the representable range, providing predictable granularity.
    \item Numbers that cannot be exactly represented within the fixed precision must be rounded or truncated, introducing small errors in computation.
\end{itemize}



%************************************
\paragraph{Deterministic Arithmetic.}
%************************************
Fixed-point arithmetic performs operations deterministically:
\begin{itemize}
    \item \textbf{Shift Operations:} A left shift multiplies the number by \( 2 \), while a right shift divides it by \( 2 \), effectively adjusting the scaling factor.
    
    \item \textbf{Addition and Subtraction:} Operations are performed like integer arithmetic, and the results remain within the fixed-point format as long as overflow does not occur.
    
    \item \textbf{Multiplication:} The product of two fixed-point numbers requires adjustment to account for the implied binary point. For two numbers \( a \) and \( b \) with \( q \) fractional bits:
\[
a \cdot b = \frac{A \cdot B}{2^q},
\]
where \( A \) and \( B \) are the integer representations of \( a \) and \( b \).

Example of $6.25 \times 0.5$ with $q=4$ fractional bits,
\begin{enumerate}

    \item Encode the operands in fixed point
    \[
        6.25_{10} = 0110.0100_2 \quad \text{ and } \quad 0.5_{10} = 0000.1000_2    
    \]
    
    \item Perform 2's complement multiplication:
    \[
    01100100_2 \cdot 00001000_2 = 0000001100100000_2.
    \]

    \item Shift the result right by $q=4$ bits and retain the lower $n=8$ bits:
    \[
    0000001100100000_2 \rightarrow 000000110010xxxx_2 \rightarrow 00110010_2
    \]

    \item Interpret the number with radix point at $q=4$:
    \[
    0011.0010_2
    \]
    This corresponds to \( 3.125 \) in decimal.
\end{enumerate}

The same example with (\( q = 3 \))
\begin{enumerate}
    \item Encode the operands in fixed-point representation:
    Represent each operand with \( q = 3 \) fractional bits:
    \[
    6.25_{10} = 00110.010_2 \quad \text{ and } \quad 0.5_{10} = 00000.100_2.
    \]

    \item Perform 2's complement multiplication:
    \[
    00110010_2 \cdot 00000.100_2 = 0000000011001000_2.
    \]

   \item Shift the result right by $q=3$ bits and retain the lower $n=8$ bits:
    \[
    0000000011001000_2 \rightarrow 0000000011001xxx_2 \rightarrow 00011001_2
    \]

    \item Interpret the number with radix point at $q=3$:
    \[
   00011.001_2
    \]
    This corresponds to \( 3.125 \) in decimal.

\end{enumerate}
    \item \textbf{Division:} Division requires rescaling to maintain precision, often introducing rounding errors:
\[
a / b = \frac{A \cdot 2^q}{B}.
\]
\end{itemize}


%*******************************************
\paragraph{Overflow and Underflow Behavior.}
%*******************************************
Unlike floating-point arithmetic, fixed-point numbers cannot dynamically adjust their range, resulting in strict overflow and underflow behavior:
\begin{itemize}
    \item Overflow: Values exceeding the maximum representable range wrap around (in unsigned systems) or saturate (in signed systems with clipping). 
    \begin{itemize}[label={}]
        \item For example, in a signed 8-bit system with 4 fractional bits:
\[
+8.0 \quad \text{is unrepresentable and results in overflow.}
\]
    \end{itemize}
    
    \item Underflow: Values closer to zero than the smallest representable value are rounded to zero, losing precision.  
    
    For \( q = 4 \), the smallest representable nonzero value is \( \pm \frac{1}{16} \).

\end{itemize}

%***********************************
\paragraph{Rounding and Truncation.}
%***********************************
Fixed-point arithmetic introduces rounding or truncation when numbers cannot be exactly represented. For example, \( 0.1 \) in decimal is approximately \( 0000.1100_2 \) in binary, resulting in a small rounding error. Common methods to deal with this include:
\begin{itemize}
    \item Truncation: Simply removes excess fractional bits, leading to predictable but potentially biased errors.
    \item Rounding: Rounds to the nearest representable value, reducing bias but requiring additional computation.

\end{itemize}

%**************************************
\subsection{Fixed Point Considerations}
%**************************************
\paragraph{Trade-offs Between Precision and Range.}
%**************************************************
Fixed-point representation inherently balances precision and range:
\begin{itemize}
    \item Increasing the number of fractional bits (\( q \)) improves precision but reduces the representable range.
    \item Increasing the number of integer bits (\( p \)) extends the range but reduces precision.
\end{itemize}
For example 
a 16-bit fixed-point number with 12 fractional bits has a higher precision but a smaller range than one with 8 fractional bits.

%*********************************************
\paragraph{Efficient Hardware Implementation.}
%*********************************************
Fixed-point arithmetic often uses the same hardware as two's complement integer arithmetic, with the distinction lying primarily in how the bits are interpreted. This close relationship allows fixed-point arithmetic to reuse existing hardware while supporting fractional precision. Arithmetic operations such as addition, subtraction, and shifts directly map to standard integer operations, reducing computational complexity.

 Fixed-point numbers are essentially integers with an \textit{implied binary point} that is not physically stored in the hardware. For example, in a 16-bit fixed-point number with 4 fractional bits, the value \( 0110.0100_2 \) is treated as \( 6.25 \) but stored and manipulated as the integer \( 100 \). Arithmetic operations like addition and subtraction can use the same integer-based hardware because the binary point’s location is an interpretation applied at the software or algorithm level.

Fixed-point numbers require consistent scaling factors. When performing operations, both operands must be in the same fixed-point format (same number of fractional bits). Multiplication results may need to be scaled back to the original format, which involves shifting the binary point. This introduces precision management, as the fractional part can lead to truncation or rounding errors.

In two's complement, the integer result is directly usable. In fixed-point arithmetic, the integer result must be interpreted based on the scaling factor to obtain the actual numerical value.




%*********************************************************
\subsection{Comparison with Floating-Point Representation}
%*********************************************************
While fixed-point arithmetic is simpler and more efficient, it lacks the flexibility and dynamic range of floating-point numbers. Floating-point systems dynamically adjust the scaling factor using an exponent, making them suitable for applications requiring high precision or a wide dynamic range.

Fixed-Point Advantages:
        \begin{itemize}
            \item Lower computational cost.
            \item Predictable rounding behavior.
            \item Simpler hardware implementation.
        \end{itemize}
        
Fixed-Point Disadvantages:
        \begin{itemize}
            \item Limited dynamic range.
            \item Susceptible to overflow and underflow.
            \item Loss of precision in arithmetic operations.
        \end{itemize}




%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Examples of Finite Real Numbers}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Practical examples of finite real numbers include:
\begin{itemize}
    \item Rational Numbers: Numbers like \( 0.75 \) or \( 0.333... \) are approximated with a finite number of decimal or binary digits, depending on the precision.
    \item Irrational Numbers: Irrationals like \( \pi \) or \( \sqrt{2} \) are represented by truncated or rounded approximations. For instance, \( \pi \) may be approximated as \( 3.14159265358979 \) in double precision.
    \item Large or Small Magnitudes: Real numbers with large or small magnitudes, such as \( 6.022 \times 10^{23} \) (Avogadro's number) or \( 1.602 \times 10^{-19} \) (charge of an electron), are stored in scientific notation using floating-point systems.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Challenges in Finite Real Representations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The finite representation of real numbers introduces challenges in computation:

\paragraph{Rounding Errors.}
Due to limited precision, some numbers cannot be represented exactly, leading to rounding errors. For instance, the decimal \( 0.1 \) has no exact binary representation and is approximated, resulting in errors in calculations.

\paragraph{Overflow and Underflow.}
Numbers exceeding the range of the floating-point format result in overflow, while numbers too close to zero result in underflow, both leading to inaccuracies or loss of data.

\paragraph{Loss of Associativity.}
Finite precision arithmetic breaks mathematical properties like associativity. For example:

\[
(a + b) + c \neq a + (b + c),
\]

when \( a, b, c \in \mathbb{R} \) are subject to rounding.



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Alternative Finite Number Representations.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Beyond traditional floating-point formats, alternative finite number representations have been developed to address specific computational challenges. 

\paragraph{Interval arithmetic} represents numbers as intervals \( [a, b] \) rather than single values, ensuring that rounding errors and uncertainties are contained within computations \cite{moore2009introduction}. For example, adding two intervals produces a new interval that encloses all possible results. This method is important in numerical analysis, optimization, and scientific computing but requires careful management to prevent interval overgrowth. 

\paragraph{Logarithmic number systems (LNS)} represent numbers as logarithms of their absolute values, offering efficient multiplication and division but complicating addition and subtraction \cite{naylor2013applications}. 

\paragraph{Tapered floating-point} formats dynamically adjust the precision of the significand based on the exponent, improving adaptability for applications requiring variable precision \cite{gustafson2015tapered}. 

\paragraph{Brain Floating Point} In machine learning, bfloat16 balances the range of `float32` with reduced precision, optimizing performance on large-scale AI workloads \cite{wang2019bfloat16}. 





