In this chapter, we make a compromise and provide an introduction to the field of digital circuits that introduces only the necessary concepts to understand the principles on which the organization and architecture of computers are based. We generally present functional or behavioral designs in preference to highly optimized structural components.

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Boolean Algebra}
%%%%%%%%%%%%%%%%%%%%%%%%%%
Boolean algebra is a branch of mathematics that deals with binary variables and logical operations. It serves as the foundation for modern digital logic and computer science, enabling the design and analysis of digital circuits and computational algorithms. Developed in the mid-19th century, Boolean algebra forms the basis of everything from simple electronic gates to complex processors.

%*********************************
\subsection{Historical Background}
%*********************************
Boolean algebra was introduced by George Boole in his seminal work, \emph{An Investigation of the Laws of Thought}, published in 1854 \cite{boole1854book}. Boole sought to formalize the process of human reasoning using a symbolic language, representing logical relationships through algebraic structures. Although initially developed as a tool for philosophy and logic, Boolean algebra found practical applications in the early 20th century when Claude Shannon demonstrated its utility in designing electrical circuits \cite{shannon38}.

Shannon's master's thesis, \emph{A Symbolic Analysis of Relay and Switching Circuits}, linked Boolean algebra with electrical switches, showing how logical operations could be implemented using relay circuits. This work laid the foundation for digital circuit design and the development of modern computers \cite{shannon38}.

%***********************
\subsection{Basic Concepts}
%************************
Boolean algebra operates on a binary set \( B = \{0, 1\} \), where:
\begin{itemize}
    \item \( 0 \) represents \textbf{false}.
    \item \( 1 \) represents \textbf{true}.
\end{itemize}

%*******************
\subsection{Operations}
%*******************
\paragraph{AND (\( \land \)).} A binary operation where \( A \land B = 1 \) if and only if both \( A = 1 \) and \( B = 1 \).
\paragraph{ OR (\( \lor \)).}  A binary operation where \( A \lor B = 1 \) if at least one of \( A \) or \( B \) equals \( 1 \).

\paragraph{NOT (\( \neg \)).} A unary operation where \( \neg A = 1 \) if \( A = 0 \), and \( \neg A = 0 \) if \( A = 1 \).

%*****************
\subsection{Laws}
%*****************
These operations obey the following basic laws:
\paragraph{Commutative Law:} 
\[
A \lor B = B \lor A, \quad A \land B = B \land A
\]

\paragraph{Associative Law:} 
\[
A \lor (B \lor C) = (A \lor B) \lor C, \quad A \land (B \land C) = (A \land B) \land C
\]

\paragraph{Distributive Law:} 
\[
A \land (B \lor C) = (A \land B) \lor (A \land C)
\]

\paragraph{Identity Law:} 
\[
A \lor 0 = A, \quad A \land 1 = A
\]

\paragraph{Complement Law:} 
\[
A \lor \neg A = 1, \quad A \land \neg A = 0
\]

%***************************
\subsection{DeMorgan's Laws}
%***************************
DeMorgan's Laws are fundamental principles in Boolean algebra, describing the relationships between logical conjunction (AND), disjunction (OR), and negation (NOT). They provide a way to express the complement of a compound expression in terms of the complements of its components. These laws are especially useful in digital logic design and simplification, as they allow for the transformation of logic expressions to equivalent forms using fewer gates or alternate gate configurations \cite{rosen2018, tanenbaum2013}.

Mathematically, DeMorgan's Laws state:
\[
\neg (A \lor B) = (\neg A) \land (\neg B), \quad \neg (A \land B) = (\neg A) \lor (\neg B)
\]

These laws can be proven using truth tables or algebraic manipulation and hold for any Boolean variables \( A \) and \( B \). In words:
\begin{itemize}
    \item The complement of a disjunction (\( A \lor B \)) is equivalent to the conjunction of the complements (\( \neg A \land \neg B \)).
    \item The complement of a conjunction (\( A \land B \)) is equivalent to the disjunction of the complements (\( \neg A \lor \neg B \)).
\end{itemize}

%**********************************************
\subsubsection{Applications of DeMorgan's Laws}
%**********************************************
DeMorgan’s Laws are widely applied in the simplification and design of digital circuits. For example, consider the simplification of the following expression that enables the implementation of the logic using AND gates and inverters rather than a more complex combination of gates \cite{rosen2018, karnaugh1953}.
\[
\neg (A \lor B \lor C) = \neg A \land \neg B \land \neg C
\]

%******************************************
\subsection{Applications in Digital Logic}
%*****************************************
Boolean algebra provides the mathematical framework for designing and analyzing digital circuits. Logical gates such as AND, OR, and NOT are implemented in hardware to perform the basic operations of Boolean algebra. More complex circuits, such as multiplexers, decoders, and arithmetic logic units (ALUs), are built using combinations of these gates.

%*************************************************
\subsubsection{Simplification of Logic Expressions}
%*************************************************
One of the uses of Boolean algebra in digital logic is the simplification of logic expressions to minimize the number of gates required in a circuit. Simplification methods include:

\begin{itemize}
    \item \textbf{Boolean identities}: Using the laws of Boolean algebra to rewrite expressions  to convert between AND-OR logic and NAND-NOR logic \cite{rosen2018}.
    \item \textbf{Karnaugh maps}: A graphical method for simplifying logic functions \cite{karnaugh1953}.
    \item \textbf{Quine-McCluskey algorithm}: A tabular approach to minimize logic expressions \cite{quine1952}.
\end{itemize}

%***************************
\subsubsection{Truth Tables - MOVE SOMEWHERE}
%***************************
A truth table is a systematic way to list all possible combinations of input values and their corresponding output values for a given Boolean expression. For example, the truth table for the XOR operation (\( A \oplus B \)) is as follows:
\[
\begin{array}{|c|c|c|}
\hline
A & B & A \oplus B \\
\hline
0 & 0 & 0 \\
0 & 1 & 1 \\
1 & 0 & 1 \\
1 & 1 & 0 \\
\hline
\end{array}
\]


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%  Digital Domain
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Digital Domain}

\subsection{Signals in Digital Domain}

In the electronic digital domain we work with two values only (see Figure \ref{fig1}):

\placedrawing[H]{figures/01_01_digital_0_and_1.lp}{The two levels of the signal in the digital domain. Low level (0 Volt) for {\tt 0} or {\tt false}, and high level ($V_{DD}$ Volt) for {\tt 1} or {\tt true}.}{fig1}

\begin{itemize}
\item 0, represented by the electrical value 0 V, having two meanings:
\begin{itemize}
\item the numerical value {\tt 0}
\item the logic value {\tt false}
\end{itemize}
\item 1, represented by the electrical value $V_{DD}$ V, having two meanings:
\begin{itemize}
\item the numerical value {\tt 1}
\item the logic value {\tt true}
\end{itemize}
\end{itemize}

The signals used in a digital system are of two types:
\begin{itemize}
\item non-periodic, with a certain structure ("random") adapted to the process of simulating the operation of a digital system
\item periods, usually with a strictly alternating structure, used as clock signals by which the operation of a digital system is synchronized.
\end{itemize}

Any simulation is based on the generation of waveforms with which the simulated circuit is stimulated on its inputs.
In the following System Verilog module, two waveforms are generated, one "random" and one usable as a periodic signal.

\begin{shaded}
\includecode{.}{waveFormGenerator}
\end{shaded}


By simulating the {\tt waveFormGenerator.sv} module, the behavior represented in Figure \ref{wave} results.



\placedrawing[H]{figures/01_02_waveSim.lp}{Waveforms generated for a Vivado System Verilog simulation}{wave}


In a digital system we can distinguish two extreme types of circuits:
\begin{itemize}
\item purely logical circuits that operate on some binary signals with some symbolic meaning (example: the transcoder for displaying the numbers from 0 to 9 using 7 segments, represented in Figure \ref{fig2}a)

\item purely numerical circuits that operate on binary signals with numerical significance (example: a modulo 16 adder for two numbers represented by 4 bits each, represented in Figure \ref{fig2}b)
\end{itemize}
Obviously, an intermediate version is always possible (example: a circuit that compares two numbers, A and B, and indicates through three logical signals of one bit each: $A>B, \; A=B, \; A\leq B$)

\placedrawing[H]{figures/01_03_7seg_and_adder.lp}{The version of digital circuits. {\bf a.} Logic circuit: trans-coder for seven-segment display. {\bf b.} Numeric circuit: four-bit numbers adder.}{fig2}



\subsection{Behavioral vs. Structural Descriptions}

The description of a digital circuit can be done in two ways:
\begin{itemize}
\item  by describing the functional behavior
\item  by describing the internal structure of the circuit
\end{itemize}
We exemplify the two methods in the case of the addition function.


\begin{exemplu}\label{adder}
Let us use the {\em System Verilog} HDL to describe an adder for 4-bit
numbers (see Figure \ref{first_ex}a). The description which
follows is a {\bf behavioral} one, because we know {\bf what} we
intend to design, but we do not know yet {\bf how} to design the
internal structure of an adder.

\placedrawing[h]{figures/01_04_4bit_adder.lp}{{\bf The first examples of  digital
systems.} {\small {\bf a.} The two 4-bit numbers adder, called
{\tt adder}. {\bf b.} The structure of an adder for 3 4-bit
numbers, called {\tt threeAdder}.} }{first_ex}

The Verilog code describing the module {\tt adder} is:
{\small
{\em
\begin{shaded*}
\begin{lstlisting}[language=Verilog, caption= , morekeywords={always_ff, always_comb, logic, loop, repeat, doInParallel, for}]
/*************************************************************************
File name:      adder.sv
Circuit name:   Adder
Description:    The module 'adder' has 2 4-bit inputs and one 4-bit output
                The circuit adds modulo 16; do not use or provide carry
                signal
*************************************************************************/
module adder(   output logic [3:0] out ,    // 4-bit output
                input  logic [3:0] in0 ,    // 4-bit input
                input  logic [3:0] in1 );   // 4-bit input

    assign out = in0 + in1;
endmodule
\end{lstlisting}
\end{shaded*}
}
}

The story just told by the previous Verilog module is: ``the 4-bit adder has two inputs, {\tt in0, in1}, one output, {\tt out}, and
its output is continuously assigned to the value obtained by
adding modulo 16 the two input numbers".

  $\diamond$
\end{exemplu}

What we just learned from the previous first simple example is
summarized in the following {\bf SystemVerilogSummary}.

\begin{verisum}{\em :

\hspace{-0.8cm} \rule[5mm]{16cm}{0.25mm} \vspace{-0.7cm}

\begin{description}
\item[/*]: begin comment
\item[*/]: end comment
\item[module]: keyword which indicates the beginning of the
description of a circuit as a module having the name which
immediately follows (in our example, the name is: {\tt adder})

\item[endmodule]: keyword which indicates the end of the module's
description which started with the previous keyword {\bf module}

\item[output]: keyword used to declare a terminal as an output (in
our example the terminal {\tt out} is declared as output)

\item[input]: keyword used to declare the terminal as an input (in
our example the terminals {\tt in0} and {\tt in1} are declared as
inputs)

\item[logic]: a data type with 4-state bits: {\tt 0, 1, x, z} where 0 means low, 1 means high, x means unknown, and z means an undriven net.

\item[initial]: a block which starts at the beginning of simulation

\item[begin ... end]: block delimiters

\item[\#]: indicate a raw value

\item[forever]: indicate an unending operation

\item[$\sim$]: negation operator

\item[assign]: keyword called the {\em continuous assignment},
used here to specify the function performed by the module (the
output {\tt out} takes continuously the value computed by adding
the two input numbers)

\item[(...)]: delimiters used to delimit the list of terminals
(external connections)

\item[,]: delimiter to separate each terminal within a list of
terminals

\item[;]: delimiter for end of line

\item[[ $ \dots$ ]]: delimiters which contains the definition of
the bits associated with a connection, for example [3:0] define
the number of bits for the three connections in the previous
example

\item[+]: the operator {\tt add}, the only one used in the
previous example.
\end{description}
\rule[5mm]{16cm}{0.25mm}
}
\end{verisum}

The description of a digital system is a {\bf hierarchical
construct} starting from a top module populated by modules, which
are similarly defined. The process continues until very simple
module are directly described. Thus, the functions $f$ and $g$ are specified by HDL programs
(in our case, in the previous example by a Verilog program).

The main characteristic of the digital design is {\bf modularity}. A
problem is decomposed in many simpler problems, which are solved
similarly, and so on until very simple problems are identified.
Modularity means also to define as many as possible identical
modules in each design. This allow to replicate many times the
same module, already designed and validated. {\em Many \& simple}
modules! Is the main slogan of the digital designer. Let's take
another example which uses as module the one just defined in the
previous example.

\begin{exemplu}
The previously exemplified module ({\tt adder}) will be used to
design a modulo 16 3-number adder, called {\tt threeAdder} (see Figure
\ref{first_ex}b). It adds 3 4-bit numbers providing a 4-bit result
(modulo 16 sum). Follows the {\bf structural} description:
{\small
{\em
\begin{shaded*}
\begin{lstlisting}[language=Verilog, caption= , morekeywords={always_ff, always_comb, logic, loop, repeat, doInParallel, for}]
/*************************************************************************
File name:      threeAdder.sv
Circuit name:   Three Input Adder
Description:    The module 'threeAdder' has 3 4-bit inputs and one 4-bit
                output. The circuit adds modulo 16 three numbers;
                do not provide carry output
*************************************************************************/
module threeAdder( output   logic [3:0]  out,
                   input    logic [3:0]  in0,
                   input    logic [3:0]  in1,
                   input    logic [3:0]  in2);

    logic [3:0] sum ;

    adder   inAdder(.out(sum),
                    .in0(in1),
                    .in1(in2)),
            outAdder(.out(out),
                     .in0(in0),
                     .in1(sum));
endmodule
\end{lstlisting}
\end{shaded*}
}
}

Two modules of {\tt adder} type (defined in the previous example)
are instantiated as {\tt inAdder}, {\tt outAdder}, they are
interconnected using the wire {\tt sum}, and are connected to the
terminals of the {\tt threeAdder} module. The resulting structure
computes the sum of three numbers. $\diamond$
\end{exemplu}

\begin{verisum}{\em :

\hspace{-0.8cm} \rule[5mm]{16cm}{0.25mm} \vspace{-0.7cm}

\begin{itemize}
\item How a previously defined module (in our example: {\tt
adder}) is two times instantiated using two different names ({\tt
inAdder} and {\tt outAdder} in our example)

\item A ``safe" way to allocate the terminals for a module
previously defined and instantiated inside the current module:
each original terminal name is preceded by a dot, and followed by
a parenthesis containing the name of the wire or of the terminal
where it is connected (in our example, {\tt outAdder( ...
.in1(sum))} means: the terminal {\tt in1} of the instance {\tt
outAdder} is connected to the wire {\tt sum})

\item The successive instantiations of the same module can be
separated by a ``,".
\end{itemize}
\rule[5mm]{16cm}{0.25mm} }
\end{verisum}

While the module {\tt adder} is a {\em behavioral} description,
the module {\tt threeAdder} is a {\em structural} one. The first
tells us {\em what} is the function of the module, and the second
tells us {\em how} its functionality is performed by using a
structure containing two instantiation of a previously defined
subsystems, and an internal connection.



\section{One-input combinational circuits}

\subsection{Formal Description}

{\small
\begin{table}[H]
  \begin{center}
    \caption{One-input circuits.}
    \label{tab:one_input_circuits}
    \begin{tabular}{|c||c|c|c|c|}
      \hline
      A&  &  NOT&  BUFFER&      \\ \hline \hline
      0&  0&   1&       0& 1    \\ \hline
      1&  0&   0&       1& 1    \\ \hline
    \end{tabular}
  \end{center}
\end{table}
}

\placedrawing[H]{figures/02_01_not.lp}{Graphic representation of the NOT and BUFFER circuits.}{fig10}

{\small
\subsection{$^*$Physical Implementation of NOT Circuit}

\subsubsection{CMOS switches}

A logic gates consists in a network of interconnected switches
implemented using the two type of MOS transistors: p-MOS and
n-MOS. How behaves the two type of transistors in specific
configurations is presented in Figure \ref{0_switch}.

A switch connected to $V_{DD}$ is implemented using a p-MOS
transistor. It is represented in Figure \ref{0_switch}a {\em off}
(generating $z$, which means {\em Hi-Z}: no signal, neither 0, nor
1) and in Figure \ref{0_switch}b it is represented {\em on}
(generating 1 logic, or {\em truth}).

A switch connected to {\em ground} is implemented using a n-MOS
transistor. It is represented in Figure \ref{0_switch}c {\em off}
(generating $z$, which means {\em Hi-Z}: no signal, neither 0, nor
1) and in Figure \ref{0_switch}e it is represented {\em on}
(generating 0 logic, or {\em false}).

\placedrawing[h]{figures/02_02_transistors.lp}{{\bf Basic switches.} {\small {\bf a}.
Open switch connected to $V_{DD}$. {\bf b}. Closed switch connected to
$V_{DD}$. {\bf c}. Open switch connected to {\em ground}. {\bf d}.
Closed switch connected to {\em ground}.}}{0_switch}

A MOS transistor works very well as an {\em on-off} switch
connecting its drain to a certain potential.  A p-MOS transistor
can be used to connect its drain to a high potential when its
gates is connected to ground, and an n-MOS transistor can connect
its drain to ground if its gates is connected to a high potential.
This complementary behavior is used to build the elementary logic
circuits.

In Figure \ref{src} is presented the {\em switch-resistor-capacitor
model} (SRC). If $V_{GS} < V_{T}$ then the transistor is {\bf off}, if
$V_{GS} \geq V_{T}$ then the transistor is {\bf on}. In both cases the
input of the transistor behaves like a capacitor, the gate-source
capacitor $C_{GS}$.

When the transistor is on its drain-source resistance is:
$$R_{ON} = R_{n}\frac{L}{W}$$
where: $L$ is the channel length, $W$ is the channel width, and $R_{n}$
is the resistance per square. The length $L$ is a constant
characterizing a certain technology. For example, if $L = 0.13\mu m$
this means it is about a $0.13\mu m$ process.

The input capacitor has the value:
$$C_{GS} = \frac{\epsilon_{OX} LW}{d}.$$
The value:
$$C_{OX} = \frac{\epsilon_{OX}}{d}$$
where: $\epsilon_{OX} \approx 3.9 \epsilon_{0}$ is the permittivity of
the silicon dioxide, is  the gate-to-channel capacitance per unit area
of the MOSFET gate.

In this conditions the gate input current is:
$$i_{G} = C_{GS} \frac{dv_{GS}}{dt}$$

\placedrawing[h]{figures/02_03_mosfet.lp}{{\bf The MOSFET switch.} {\small The {\em
switch-resistor-capacitor} model consists in the two states: OF
($V_{GS} < V_{T}$), and ON ($V_{GS} \geq V_{T}$). In both states the
input is defined by the capacitor $C_{GS}$.}}{src}


\begin{exemplu}
For an AND gate with low strength, with $W = 1.8\mu m$, in $0.13
\mu m$ technology, supposing $C_{OX} = 4fF/\mu m^{2}$, results the
input capacitance: $$C_{GS} = 4 \times 0.13 \times 1.8 fF = 0.936
fF$$ Assuming $R_{n} = 5K\Omega$, results for the same gate:
$$R_{ON} = 5 \times \frac{0.13}{1.8} \; K\Omega = 361 \Omega$$

$\diamond$
\end{exemplu}

\subsubsection{The Inverter}


\paragraph{The static behavior.}
The smallest and simplest logic circuit -- the invertor -- can be built
using a pair of complementary transistors, connecting together the two
gates as input and the two drains as output, while the n-MOS source is
connected to ground (interpreted as logic 0) and the p-MOS source to
$V_{DD}$ (interpreted as logic 1). Results the circuit represented in
Figure \ref{0_inv}.

\placedrawing[h]{figures/02_04_inverter.lp}{{\bf Building an invertor.} {\small {\bf
a}. The invertor circuit. {\bf b}. The logic symbol for the invertor
circuit.}}{0_inv}

The behavior of the invertor consist in combining the behaviors of the
two switches previously defined. For $in=0$ pMOS is $on$ and nMOS is
$off$ the output generating $V_{DD}$ which means 1. For $in=1$ pMOS is
$off$ and nMOS is $on$ the output generating 0.

The static behavior of the inverter (or NOT) circuit can be easy
explained starting from the switches described in Figure
\ref{0_switch}. Connecting together a switch generating $z$ with a
switch generating 1 or 0, the connection point will generate 0 or
1.

\paragraph{Dynamic behavior.}
The propagation time of an inverter can be analyzed using the two
serially connected invertors represented in Figure \ref{switch}. The
delay of the first invertor is generated by its capacitive load,
$C_{L}$, composed by:
\begin{itemize}
\item its parasitic drain/bulk capacitance, $C_{DB}$, is the intrinsic  output capacitance of the first invertor
\item wiring capacitance, $C_{wire}$, which depends on the length of the wire (of width $W_w$ and of length $L_w$) connected between the two invertors: $$C_{wire} = C_{thickox} W_w L_w$$
\item next stage input capacitance, $C_{G}$, approximated by summing the gate capacitance for pMOC and nMOS transistors: $$C_G = C_{Gp} + C_{Gn} = C_{ox}(W_p L_p + W_n L_n)$$
\end{itemize}

\placedrawing[h]{figures/02_05_propagation_time.lp}{{\bf The propagation time.} {\small
}}{switch}

The total load capacitance
$$C_L = C_{DB} + C_{wire} + C_G$$
is sometimes dominated by $C_{wire}$. For short connections $C_G$ dominates, while for big {\em fan-out}\index{fan-out: the number on input gates  connected to the output of a digital circuit} both, $C_{wire}$ and $C_G$ must be considered.

The signal $V_A$ is used to measure the propagation time of the first NOT in Figure \ref{switch}a. It is generated by an ideal pulse generator with output impedance 0. Thus, the rising time and the falling time of this signal are considered 0 (the input capacitance of the NOT circuit is charged  or discharged in no time).

The two delay times (see Figure \ref{switch}c) associated to an invertor
(to a gate in the general case) are defined as follows:
\begin{itemize}
    \item $t_{pLH}$: the time interval between the moment the input
    switches in 0 and the output reaches $V_{OH}/2$ coming from 0
    \item $t_{pHL}$: the time interval between the moment the input
    switches in 1 and the output reaches $V_{OH}/2$ coming from $V_{OH}$
\end{itemize}

Let us consider the transition of $V_A$ from 0 to $V_{OH}$ at $t_r$ (rise edge). Before transition, at $t_r^-$, $C_L$ is fully charged and $V_B = V_{OH}$. In Figure \ref{switch}b is represented the equivalent circuit at $t_r^+$, when pMOS is off and nMOS is on. In this moment starts the process of discharging  the capacitance $C_L$ at the constant current
$$I_{Dn(sat)} = \frac{1}{2} \mu_n C_{ox} \frac{W_n}{L_n} (V_{OH} - V_{Tn})^2$$
In Figure \ref{switch1}, at $t_r^-$ the transistor is cut, $I_{Dn} = 0$. At $t_r^+$ the  nMOS transistor switch in saturation and becomes an ideal {\em constant current generator} which starts to discharge $C_L$ linearly at the constant current $I_{Dn(sat)}$. The process continue until $V_{OUT} = V_{OH}$, according to the definition of $t_{pHL}$.


\placedrawing[h]{figures/02_06_nmos_output.lp}{{\bf The output characteristic of the nMOS transistor.} {\small
}}{switch1}

In order to compute $t_{pHL}$ we take into consideration the constant value of the discharging current which provide a linear variation of $v_{OUT}$.
$$\frac{dv_{out}}{dt} = \frac{d}{dt}(\frac{q_L}{C_L}) = \frac{-I_{Dn(sat)}}{C_L}$$
$$\frac{dv_{out}}{dt} = \frac{\frac{V_{OH}}{2} - V_{OH}}{t_{pHL}}$$
We solve the equations for $t_{pHL}$:
$$t_{pHL} = C_L \frac{1}{\mu_n C_{ox} \frac{W_n}{L_n} (V_{OH} - V_{Tn})} \frac{V_{OH}}{V_{OH} - V_{Tn}}$$
Because:
$$R_{ONn} = \frac{1}{\mu_n C_{ox} \frac{W_n}{L_n} (V_{OH} - V_{Tn})}$$
results:
$$t_{pHL} = C_L R_{ONn} \frac{1}{1 - \frac{V_{Tn}}{V_{OH}}} = k_n R_{ONn} C_L = k_n \tau_{nL} $$
where:
\begin{itemize}
\item $\tau_{nL}$ is the {\em constant time} associated to the H-L transition
\item $k_n$ is a constant associated to the technology we use; it goes down when $V_{OH}$ increases or $V_T$ decreases
\end{itemize}

The speed of a gate depends by its dimension and by the capacitive load
it drives. For a big $W$ the value of $R_{ON}$ is small charging or
discharging $C_{L}$ faster.

For $t_{pLH}$ the approach is similar. Results: $t_{pLH} = k_p \tau_{pL}$.

By definition the propagation time associated to a circuit is: $$t_p = (t_{pLH} + t_{pHL})/2$$
its value being dominated by the value of $C_L$ and the size (width) of the two transistors, $W_n$ and $W_p$.
}

\section{Two-input gates}

\subsection{Formal Description}

{\small
\begin{table}[H]
  \begin{center}
    \caption{Two-input gates.}
    \label{tab:two_input_gates}
    \begin{tabular}{|c|c||c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|}
      \hline
      A& B& &  NOR&  & &  & & XOR& NAND& AND& NXOR&  &  &  &  & OR& \\ \hline \hline
      0& 0& 0&   1& 0&     1& 0&     1&   0&    1&   0&    1& 0& 1& 0& 1&  0& 1\\ \hline
      0& 1& 0&   0& 1&     1& 0&     0&   1&    1&   0&    0& 1& 1& 0& 0&  1& 1\\ \hline
      1& 0& 0&   0& 0&     0& 1&     1&   1&    1&   0&    0& 0& 0& 1& 1&  1& 1\\ \hline
      1& 1& 0&   0& 0&     0& 0&     0&   0&    0&   1&    1& 1& 1& 1& 1&  1& 1\\ \hline
    \end{tabular}
  \end{center}
\end{table}
}

The currently used two-input gates are (see Tab. \ref{tab:two_input_gates}):
\begin{description}
\item[AND]: A$\cdot$B = AB = 1 only if both A and B are 1, else AB = 0\\
    Numerical interpretation: arithmetic product \\
    Symbolic interpretation: {\bf gate} because A=1 opens the way for B
\item[NAND]: (AB)' = 0 only if both A and B are 1, else AB = 1
\item[OR]: A+B = only if at least one of A and B is 1, else A+B = 0
\item[NOR]: (A+B) = 0 only if at least one of A and B is 1, else A+B = 1
\item[XOR]: exclusive OR, because it is excluded the case A=B=1 in determining 1 on the output\\
    A$\oplus$B = 1 when only one of the input, A or B, is one\\
    Logic interpretation: conditioned inverter, i.e., {\em if} A = 0 {\bf then}  A$\oplus$B = B {\bf else}  A$\oplus$B = B'
    Numerical interpretation: modulo-2 sum\\
    Symbolic interpretation: anticoincidence
\item[NXOR]: (A$\oplus$B')\\
    Symbolic interpretation: coincidence
\end{description}

\placedrawing[H]{figures/02_07_two_input_gates.lp}{Graphical representation of the most used two-input gates.}{fig11}
\placedrawing[H]{figures/02_08_non_inverting_gate_waveforms.lp}{Wave forms for the non-inverting gates.}{fig16}


In Figure \ref{fig16} is represented the behavior of the main gates, AND, OR and XOR, driven by the signal A snd B represented in the first two wave forms. Besides the logic response to the inputs A and B there are represented the main time characteristics:
\begin{itemize}
    \item $t_+$: the positive transition time (the duration of the positive edge)
    \item $t_-$: the negative transition time (the duration of the negative edge)
    \item $t_{pLH}$: the propagation time  of the signal through the gate for the transition from Low to High
    \item $t_{pHL}$: the propagation time  of the signal through the gate for the transition from High to Low
\end{itemize}



All the previous times are different from positive and negative transition and from a type of gate to another. For our approach differences are small and unimportant.




\section{Many-input gates}

The number $N$ of $n$-input gates is: $N^{2^n}$, because the number of input binary configuration is $2^n$. Therefore, they cannot be listed. For $n>2$ the Boolean functions are managed using Boolean algebra rules.

Let's consider some significant examples:

\begin{itemize}
    \item (A')' = A
    \item AB + AB' = A(B + B') = A $\cdot$ 1 = A\\
    {\em because, when B=1 the expression takes the value A, and when B=0 the expression still takes the value A, we conclude that the value of B does not matter and the expression takes the value A}.
    \item A + A'B = A + B\\
    to prove it the method of truth table is used, as follow:
    \begin{table}[H]
  \begin{center}
    \caption{Example of proof using the truth table.}
    \label{tab:table1}
    \begin{tabular}{|c|c||c|c|c||c|}
      \hline
      A& B&  A'& A'B& A+A'B& A+B   \\ \hline \hline
      0& 0&   1&   0&     0&   0   \\ \hline
      0& 1&   1&   1&     1&   1   \\ \hline
      1& 0&   0&   0&     1&   1   \\ \hline
      1& 1&   0&   0&     1&   1   \\ \hline
    \end{tabular}
  \end{center}
\end{table}
    \item A $\oplus$ B = (A' $\oplus$ B)' = (A $\oplus$ B')' = A' $\oplus$ B'
    \item de Morgan rule: A + B = (A'B')' \\
    {\em A or B is 1 is equivalent to the fact that it is not true that A is 0 and B is 0}.
    \item de Morgan rule: AB = (A' + B')'\\
    {\em A and B are 1 is equivalent to the fact that it is not true that A is 0 or B is 0.}

\end{itemize}



Fortunately for us, synthesis software tools (Hardware Description Languages (HDLs) such as Verilog or VHDL) "know" how to use Boolean algebra very efficiently, saving us the effort of minimizing combinational logic circuits.

\subsection{Seven-Segment Display}

A seven-segment display is driven by a trans-coder which receives four-bit coded decimal numbers, from 0 to 9, and convert them in 7-bit codes, a to g, indicating the segments to be activated, according to Table \ref{sevenSeg}.

The logic expression for the output {\tt a} is:
\begin{center}
$a =   B'_3 \cdot B'_2 \cdot B'_1\cdot B'_0  + B'_3 \cdot B'_2 \cdot B_1\cdot B'_0  +
        B'_3 \cdot B'_2 \cdot B_1\cdot B_0  + B'_3 \cdot B_2 \cdot B'_1\cdot B_0  + $ \\
        $B'_3 \cdot B_2 \cdot B_1\cdot B'_0  + B'_3 \cdot B_2 \cdot B_1\cdot B_0  +
        B_3 \cdot B'_2 \cdot B'_1\cdot B'_0  + B_3 \cdot B'_2 \cdot B'_1\cdot B_0  $
\end{center}
or a simpler version by negatin the negated function:
$$a =  (a')' =  (B'_3 \cdot B'_2 \cdot B'_1\cdot B_0  + B'_3 \cdot B_2 \cdot B'_1\cdot B'_0)' = (B'_3 \cdot B'_1 \cdot (B_2 \oplus B_0))' = B_3 + B_1 + (B_2 \oplus B_0)'$$

\begin{table}[H]
  \begin{center}
    \caption{Seven-segment display.}
    \label{sevenSeg}
    \begin{tabular}{|c|c|c|c||c|c|c|c|c|c|c|}
      \hline
     $B_3$ & $B_2$& $B_1$& $B_0$& a& b& c& d& e& f& g \\ \hline \hline
      0& 0& 0& 0& 1& 1& 1& 1& 1& 1& 0 \\ \hline
      0& 0& 0& 1& 0& 1& 1& 0& 0& 0& 0 \\ \hline
      0& 0& 1& 0& 1& & & & & & \\ \hline
      0& 0& 1& 1& 1& & & & & & \\ \hline
      0& 1& 0& 0& 0& & & & & & \\ \hline
      0& 1& 0& 1& 1& & & & & & \\ \hline
      0& 1& 1& 0& 1& & & & & & \\ \hline
      0& 1& 1& 1& 1& & & & & & \\ \hline
      1& 0& 0& 0& 1& & & & & & \\ \hline
      1& 0& 0& 1& 1& & & & & & \\ \hline
    \end{tabular}
    %\label{sevenSeg}
  \end{center}
\end{table}

The System Verilog description of the circuit is sketched in the following:
{\small
\begin{shaded*}
\begin{lstlisting}[language=Verilog, caption= , morekeywords={always_ff, always_comb, logic, loop, repeat, doInParallel, for}]
/*************************************************************************
File name:      sevenSegmDys.sv
Circuit name:   Seven-Segment Display
Description:
*************************************************************************/
module sevenSegmDys(output  logic [6:0] seg     ,
                    input   logic [3:0] number  );

    always_comb case(number)
                    4'b0000: seg = 7'b1111110   ;
                    4'b0001: seg = 7'b0110000   ;
                    4'b0010: seg = 7'b  ;
                    4'b0011: seg = 7'b  ;
                    4'b0100: seg = 7'b  ;
                    4'b0101: seg = 7'b  ;
                    4'b0110: seg = 7'b  ;
                    4'b0111: seg = 7'b  ;
                    4'b1000: seg = 7'b  ;
                    4'b1001: seg = 7'b  ;
                    default: seg = 7'b  ;
                endcase
endmodule
\end{lstlisting}
\end{shaded*}
}

\begin{verisum}{\em :

\hspace{-0.8cm} \rule[5mm]{16cm}{0.25mm} \vspace{-0.7cm}

\begin{description}
    \item[always\_comb]: defines a combinational block that describes the behavior of the output of the circuit as a combination based on the value of the input variables. The blocking assignment, {\tt =}, is used to assure the combinational behavior.
    \item[4'b0010]: instantiate a variable as a 4-bit number
    \item[case]: selects a block of code to be executed based on the value of a given signal in our design. Once a match is found for the input signal value, the action associated with that value will be executed
    \item[default]: if no condition is fulfilled, the default action is selected for execution.
\end{description}
\rule[5mm]{16cm}{0.25mm} }
\end{verisum}

\section{Generic combinational circuits}

\subsection{Decoder}

Decoding means to identify with a one bit signal each binary code applied to the input of the circuit. Therefor, a $n$-bit input code applied on the input of a decoder implies a number of $2^n$ outputs. Only one output is active at a time corresponding to the input code. Thus, for the decoder represented in Figure \ref{fig14}, if $\{x_{n-1}, \ldots, x_0\} = 00 \ldots 0101$, then $y_5=1$ and all the other outputs are 0.

\placedrawing[H]{figures/02_09_decoder.lp}{Decoder with $n$-bit input (DCDn).}{fig14}

The System Verilog code for the decoder circuit follows. The description is behavioral.

\begin{verisum}{\em :

\hspace{-0.8cm} \rule[5mm]{16cm}{0.25mm} \vspace{-0.7cm}

\begin{description}
    \item[$<<$]: shift left logical, i.e., multiplication with 2 to the power of the number of binary shifts
\end{description}
\rule[5mm]{16cm}{0.25mm} }
\end{verisum}

{\small
\begin{shaded*}
\begin{lstlisting}[language=Verilog, caption= , morekeywords={always_ff, always_comb, logic, loop, repeat, doInParallel, for}]
/*************************************************************************
File name:      dcd.sv
Circuit name:   4-input decoder
Description:
*************************************************************************/
module dcd( output  logic [15:0]    dcdOut  ,
            input   logic [3:0]     dcdIn   );

     assign dcdOut = 1 << dcdIn ;
endmodule
\end{lstlisting}
\end{shaded*}
}



In Figure \ref{fig8}a, is represented the smallest decoder, the elementary DCD, eDCD. Using two eDCD and 4 ANDs, a two-input decoder, DCD2, is designed (see Figure \ref{fig8}b). The rule to increase the number of inputs becomes evident with the third example: DCD3 (see Figure \ref{fig8}c).


\placedrawing[H]{figures/02_10_recursive_decoder.lp}{Recursively defined 3-input decoder (DCD). {\bf a.} Elementary decoder (eDCD). {\bf b.} Two-input decoder (DCD2). {\bf c.} Three-input DCD (DCD3).}{fig8}


\subsection{Multiplexor}

\subsection{Elementary multiplexor}

Any combinational circuit can be expressed using NAND function. But a more interesting "fundamental brick" could be a circuit corresponding to the ubiquitous {\bf {\em if-then-ellse}} statement. It is the elementary multiplexor (eMUX) represented in Figure \ref{fig9},
where:

\placedrawing[H]{figures/02_11_mux.lp}{Elementary multiplexor (eMUX). {\bf a.} The circuit structure: an eDCD serially connected with an AND-OR circuit. {\bf b.} The logic symbol.}{fig9}

\begin{center} O = {\bf if} (S) {\bf then} I1 {\bf else} I0 \end{center}

\subsection{Many-input multiplexor}

The multiplexor grabs bits from $m=2^n$ inputs selected by a $n$-bit code.

\placedrawing[H]{figures/02_12_mux_generalized.lp}{The general definition of the multiplexor circuit (MUXn). {\bf a.} The structure of MUXn: a DCDn serially connected with a $m=2^n$ ANDs on the first level of an AND-OR structure. {\bf b.} The logic symbol for MUXn.}{fig12}


\begin{verisum}{\em :

\hspace{-0.8cm} \rule[5mm]{16cm}{0.25mm} \vspace{-0.7cm}

\begin{description}
    \item[{\tt a[b]}]: selects the bit indicated by {\tt b} from the binary word {\tt a}
\end{description}
\rule[5mm]{16cm}{0.25mm} }
\end{verisum}

{\small
\begin{shaded*}
\begin{lstlisting}[language=Verilog, caption= , morekeywords={always_ff, always_comb, logic, loop, repeat, doInParallel, for}]
/*************************************************************************
File name:      mux.sv
Circuit name:   4-input multiplexor
Description:
*************************************************************************/
module mux( output  logic           muxOut  ,
            input   logic [15:0]    muxIn   ,
            input   logic [3:0]     muxSel  );

     assign muxOut = muxIn[muxSel] ;
endmodule
\end{lstlisting}
\end{shaded*}
}


\subsection{Demultiplexor}

The demultiplexer transfers the input signal X to one of the $m=2^n$ outputs, $y_0, y_1, \ldots, y_{m-1}$, indicated by the $n$-bit selection input, $x_0, x_1, \ldots, x_{n-1}$. In Figure \ref{fig13} a demultiplexer is built using a decoder which open only one of the $m$ AND gate sending the $X$ input to the selected output.

\placedrawing[H]{figures/02_13_demux.lp}{Demultiplexor as a DCD whose outputs are ANDed with the one bit signal to be distributed according to the $n$-bit input code. }{fig13}


For example: if $$\{x_{n-1} x_{n-2}, \ldots, x_1, x_0\} = 00 \ldots 011$$ then
$$\{y_{m-1} y_{m-2} \ldots y_0 y_1\} = 00 \ldots 0X000$$

The code describing behaviorally the circuit follows:

{\small
\begin{shaded*}
\begin{lstlisting}[language=Verilog, caption= , morekeywords={always_ff, always_comb, logic, loop, repeat, doInParallel, for}]
/*************************************************************************
File name:      dMux.sv
Circuit name:   4-input demultiplexor
Description:    Demultiplexor is an enabled decoder
*************************************************************************/
module dMux(output  logic [15:0]    dMuxOut     ,
            input   logic [3:0]     dMuxIn      ,
            input   logic           dMuxEnable  );

     assign dMuxOut = dMuxEnable << dMuxIn ;
endmodule
\end{lstlisting}
\end{shaded*}
}
Now, instead of left-shifting the number 1 as in the decoder, the one-bit value $X$ is left-positioned.


\section{Function oriented combinational circuits}

\subsection{Half Adder}

As we have seen, the XOR circuit calculates the sum modulo 2 but does not differentiate between 0+0 and 1+1. This differentiation involves the calculation of the arithmetical overshoot, which we denote by carry (to the next binary order). Carry signal is activated when both inputs are 1. Therefore, besides the XOR gate we must add an AND gate as in Figure \ref{fig3}. We call the resulting circuit half adder because it does not take into account the carry signal provided by the previous binary order.

\placedrawing[H]{figures/02_14_half_adder.lp}{Half adder (HA).The XOR circuit compute the {\em modulo-2} sum while the AND gate compute the carry value.}{fig3}


\subsection{Increment}

Incrementing means adding 1 to a number. The schematic for a 4-bit increment circuit, INC4, is represented in Figure \ref{fig4}, where if {\tt inc = 1} the output $out_0 <= in'_0$  and {\tt cr} of $HA_0$ if it is activated command the increment for $HA_1$, and so on.

\placedrawing[H]{figures/02_15_increment.lp}{Increment circuit for 4-bit numbers (INC4) implemented by serially connected $4$ HAs.}{fig4}

If each 2-input AND from $HA_i$ has, for example,  a propagation time, $t_{pAND2} = 1ns = 10^{-9}sec$, and the 2-input XOR has the propagation time, $t_{pXOR2} = 1.5ns$, then the total execution time for the INC4 circuit is:
$$t_{pINC4} = 3 \times t_{pAND2} + t_{pXOR2} = 4.5ns$$
Thus, when the good news that the size of the $n$-bit number increment circuit, $S_{INCn}$, is proportional with the input size, the bad news is that the execution time, i.e., depth, $D_{INCn}$, is in same magnitude order:
$$S_{INCn} \in O(n)$$
$$D_{INCn} \in O(n)$$
(The depth of the circuit can be reduced by increasing the size (but in the limits of the same magnitude order) using a prefix network.)

\subsection{Adder/subtractor}

Full adder (FA) adds two one-bit numbers with the carry provided from the previous binary position and generate the sum and the carry value for the next binary position. In Figure \ref{fig5}a, a FA is composed from two HAs and an OR circuit used to grab the carry generated by the first HA or by the second HA. The first FA adds the two input bits, A and B, while the second FA adds the value of the carry, C, generated by the previous binary order.

\placedrawing[H]{figures/02_16_adder_full.lp}{Adder. {\bf a.} Full Adder (FA). {\bf b.} 4-bit adder (ADD4).}{fig5}

\placedrawing[H]{figures/02_17_adder_subtractor.lp}{Adder/Substractor}{fig6}


The main problem to be solved, but not in this lecture, is the execution time, i.e., the maximum time interval from the moment when an input changes until the moment when all the outputs of the circuit reach the correct final value. This time interval is due to propagation through logic gates on the longest path. In our case, the longest path is the one from input C to output $C_{out}$. On this path there are 8 logic gates, one AND and one OR for each binary order. In the general case, the time is proportional to $n$ (it is in $O(n)$).





Subtract operation is performed by adding the 2s complement:

\begin{verbatim}
        +5 = 0_0101
        -5 = 1_1010 + 1 = 1_1011
        +5 = 0_0100 + 1 = 0_0101

        -5 + 2 = 1_1011 +
                 0_0010 =
                 1_1101 = -3, because 0_0010 + 1 = 0_0011 = 3
\end{verbatim}

Therefore, each $B_i$ is complemented and the increment with 1 is generated by inverting the carry input. All the inversions are performed under de command {\tt sub} applied to $5$ XORs.


\subsection{{\bf {\em Arithmetic \& Logic Unit}}}\label{alu1}

An Arithmetic \& Logic Unit (ALU) has two kinds of connections: data inputs and outputs, for the $n$-bit operands, and command inputs for the operations applied to operands (see Fig \ref{fig37}). In Figure \ref{fig15}, is represented the generic form of an ALU with 8 functions each performed in a distinct module, {\tt func0, ..., func7}, whose outputs are selected by a multiplexor as the result by the {\tt func} code. The two operands, {\tt leftOp[n-1:0], rightOp[n-1:0]}, are applied to all the 8 modules.

\placedrawing[H]{figures/02_18_alu.lp}{ALU.}{fig37}


\placedrawing[H]{figures/02_19_8function_alu.lp}{The generic form of a 8-function Arithmetic \& Logic Unit for $n$-bit words (ALUn).}{fig15}

{\small
\begin{shaded*}
\begin{lstlisting}[language=Verilog, caption= , morekeywords={always_ff, always_comb, logic, loop, repeat, doInParallel, for}]
/*************************************************************************
File name:      alu.sv
Circuit name:   arithmetic and logic unit
Description:    the circuit selects, using the selection code 'func', one
                of the 8 functions
*************************************************************************/
module ALU(input    logic         crIn          ,
            input   logic [2:0]   func          ,
            input   logic [31:0]  left, right   ,
            output  logic         crOut         ,
            output  logic [31:0]  out           );

  always_comb case(func)
                3'b000: {crOut, out} = left + right + crIn;     //add
                3'b001: {crOut, out} = left - right - crIn;     //sub
                3'b010: {crOut, out} = {1'b0, left & right};    //and
                3'b011: {crOut, out} = {1'b0, left | right};    //or
                3'b100: {crOut, out} = {1'b0, left ^ right};    //xor
                3'b101: {crOut, out} = {1'b0, ~left};           //not
                3'b110: {crOut, out} = {1'b0, left};            //left
                3'b111: {crOut, out} = {1'b0, left >> 1};       //shr
            endcase
endmodule
\end{lstlisting}
\end{shaded*}
}

\begin{verisum}{\em :

\hspace{-0.8cm} \rule[5mm]{16cm}{0.25mm} \vspace{-0.7cm}

\begin{description}
    \item[\{a,b\}]: concatenates the binary word  {\tt b} to the binary word {\tt a}
    \item[{\bf -}]:  is the operator {\tt minus}
    \item[\&]: is the operator {\tt bitwise AND}
    \item[${\bf \vert}$]: is the operator {\tt bitwise OR}
    \item[${\bf \hat{}}$]: is the operator {\tt bitwise XOR}
    \item[${\bf >>}$]: is the operator {\tt shift right logical}
\end{description}
\rule[5mm]{16cm}{0.25mm} }
\end{verisum}

Actual implementations are sometimes optimized by implementing two ore more modules sharing partially the same circuits. A small and simple example is represented in Figure \ref{fig7}, where: {\small
\begin{itemize}
\item {\tt func[2:0]} to select the function:
    \begin{description}
        \item[{\tt func = 000}] {\tt add: \{crOut, result\} = leftOp + rightOp + crIn}
        \item[{\tt func = 100}] {\tt sub: \{crOut, result\} = leftOp - rightOp - crIn}
        \item[{\tt func = 010}] {\tt and: \{crOut, result\} = \{1'b0, leftOp \& rightOp\}} (bitwise AND)
        \item[{\tt func = 001}] {\tt xor: \{crOut, result\} = \{1'b0, leftOp $\oplus$ rightOp\}} (bitwise XOR)
        \item[{\tt func = 011}] {\tt shr: \{crOut, result\} = \{leftOp[0], serialIn, leftOp[n-1:1]\}}
    \end{description}
\item {\tt leftOp[n-1:0]}
\item {\tt rightOp[n-1:0]}
\item {\tt crIn}
\item {\tt crOut}
\item {\tt serialIn}: the value loaded in the most significant position in the shift right operation, {\tt shr}, allowing the following types of shifts:
    \begin{itemize}
        \item {\tt shr: \{crOut, result\} = \{leftOp[0], 1'b0, leftOp[n-1:1]\}}
        \item {\tt ash: \{crOut, result\} = \{leftOp[0], leftOp[n-1], leftOp[n-1:1]\}}
        \item {\tt rot: \{crOut, result\} = \{leftOp[0], leftOp[0], leftOp[n-1:1]\}}
        \item {\tt csh: \{crOut, result\} = \{leftOp[0], crIn, leftOp[n-1:1]\}}
    \end{itemize}
\item {\tt result[n-1:0]}
\end{itemize}
}

\placedrawing[H]{figures/02_20_nbit_alu.lp}{A simple Arithmetic \& Logic Unit for $n$-bit words (ALUn).}{fig7}

The implementation takes into account the fact that a HA is implemented using an AND gate and a XOR gate (see Figure \ref{fig3}). Thus, in each of the $n$ slices of ALU only an adder/subtractor is implemented, a XOR is added for the subtract function and for the two logic functions the outputs of the first HA are directly used.

\section{Problems}


\subsection{Waveforms}

\begin{problem}
Generate the following waveforms:
$$W = a^3c^4ba^2$$
where: $a = 2'b01$, $b = 2'b10$, $c = 2'b11$. and each value is maintained 2 time units (\#2).

$\diamond$
\end{problem}

{\bf Solution}

Because each symbol is codded on 2 bits, results 2 wave forms: $w_1$ and $w_0$, as follows:

W =
$\begin{bmatrix}
w_1 \\
w_0
\end{bmatrix}$
=
$\begin{bmatrix}
0 & 0 & 0 & 1 & 1 & 1 & 1 & 1 & 0 & 0\\
1 & 1 & 1 & 1 & 1 & 1 & 1 & 0 & 1 & 1
\end{bmatrix}$

{\small
\begin{shaded*}
\begin{lstlisting}[language=Verilog, caption= , morekeywords={always_ff, always_comb, logic, loop, repeat, doInParallel, for}]
/*************************************************************************
File name:      waveform1.sv
Description:    W = aaaccccbaa =
*************************************************************************/
module waveform1;
    logic [1:0] W;

    parameter   a = 2'b01,
                b = 2'b10,
                c = 2'b11;

    initial begin       W = a   ;
                    #6  W = c  	;
                    #8  W = b  	;
                    #2  W = a  	;
                    #4  $stop   ;
            end
endmodule
\end{lstlisting}
\end{shaded*}
}
The resulting waveforms are captured in Figure \ref{w1}.

\placedrawing[H]{figures/02_21_waveform1.lp}{Output for waveform1.sv}{w1}


\begin{problem}
Generate the following waveforms:
$$(ab^2c)^3$$
where: $W = a = 2'b11$, $b = 2'b10$, $c = 2'b01$,  and each value is maintained 2 time units (\#4).

$\diamond$
\end{problem}


\begin{problem}
Generate a clock signal with 2 time units period and the following waveforms:
$$(a^3b^2c)^2$$
where: $W = a = 2'b11$, $b = 2'b10$, $c = 2'b01$,  and each value is changed synchronously with the negative edge of the clock signal.

$\diamond$
\end{problem}


\subsection{Combinatorial Circuits}

\begin{problem}
Design in System Verilog the structural solution at the gate level for a half adder.

$\diamond$
\end{problem}

{\bf Solution}

A half adder is defined by the following Boolean expressions:
$$sum = a \oplus b$$
$$cr = a \cdot b$$

{\small
\begin{shaded*}
\begin{lstlisting}[language=Verilog, caption= , morekeywords={always_ff, always_comb, logic, loop, repeat, doInParallel, for}]
/*************************************************************************
File name:      halfAdder.sv
Description:
*************************************************************************/
module halfAdder(   input   logic a, b,
                    output  logic sum, cr);

    xor myXor(sum, a, b);
    and myAnd(cr, a, b);
endmodule
\end{lstlisting}
\end{shaded*}
}
Using Vivado toll the elaborated design represented in Figure \ref{halfadd} is provided.

\placedrawing[H]{figures/02_22_half_adder.lp}{}{halfadd}

\begin{problem}
Design in System Verilog the structural solution at the gate level for a full-half adder using two solutions:
\begin{itemize}
    \item describe at the gate level
    \item a hierarchical approach based on the solution of the previous problem using a top module where two half-adders are instantiated.
\end{itemize}
Provide the simulation of the circuit to test exhaustively its behavior.

$\diamond$
\end{problem}

{\bf Solution}

For the first solution we have the following design:
{\small
\begin{shaded*}
\begin{lstlisting}[language=Verilog, caption= , morekeywords={always_ff, always_comb, logic, loop, repeat, doInParallel, for}]
/*************************************************************************
File name:      fullAdder.sv
Description:
*************************************************************************/
module fullAdder(   input   logic a, b, crIn,
                    output  logic sum, crOut);
    logic sum0, cr0, cr1;

    xor myXor(sum0, a, b);
    xor outXor(sum, sum0, crIn);
    and myAnd1(cr0, a, b);
    and myAnd2(cr1, sum0, crIn);
    or myOr(crOut, cr0, cr1);
endmodule
\end{lstlisting}
\end{shaded*}
}
Using Vivado toll the elaborated design represented in Figure \ref{fulladd} is provided.

\placedrawing[H]{figures/02_23_full_adder.lp}{}{fulladd}

For the second solution we have the following design:
{\small
\begin{shaded*}
\begin{lstlisting}[language=Verilog, caption= , morekeywords={always_ff, always_comb, logic, loop, repeat, doInParallel, for}]
/*************************************************************************
File name:      hFullAdder.sv
Description:    Hierarchical Full-Adder
*************************************************************************/
module hFullAdder(  input   logic a, b, crIn,
                    output  logic sum, crOut);
    logic sum0, cr0, cr1;

    halfAdder   ha0(.a  (a      ),
                    .b  (b      ),
                    .sum(sum0   ),
                    .cr (cr0    )),
                ha1(.a  (sum0   ),
                    .b  (crIn   ),
                    .sum(sum    ),
                    .cr (cr1    ));
    or outOr(crOut, cr0, cr1);
endmodule
\end{lstlisting}
\end{shaded*}
}
Using Vivado toll the elaborated design represented in Figure \ref{hfulladd} is provided.

\placedrawing[H]{figures/02_24_full_adder_from_half_adders.lp}{}{hfulladd}

For testing the functionality of the full-adder the following module is designed:

{\small
\begin{shaded*}
\begin{lstlisting}[language=Verilog, caption= , morekeywords={always_ff, always_comb, logic, loop, repeat, doInParallel, for}]
/*************************************************************************
File name:      testFullAdder.sv
Description:    Hierarchical Full-Adder
*************************************************************************/
module testFullAdder();
    logic a, b, crIn, sum, crOut;

    hFullAdder dut(a, b, crIn, sum, crOut);

    initial begin       {a,b,crIn} = 3'b000 ;
                    #1  {a,b,crIn} = 3'b001 ;
                    #1  {a,b,crIn} = 3'b010 ;
                    #1  {a,b,crIn} = 3'b011 ;
                    #1  {a,b,crIn} = 3'b100 ;
                    #1  {a,b,crIn} = 3'b101 ;
                    #1  {a,b,crIn} = 3'b110 ;
                    #1  {a,b,crIn} = 3'b111 ;
                    #1  $stop               ;
            end
endmodule
\end{lstlisting}
\end{shaded*}
}

\placedrawing[H]{figures/02_25_full_adder_waveform.lp}{}{sfulladd}

\section{Exercises}

\begin{problem}
Design a 4-bit number adder with carry input and output.

$\diamond$
\end{problem}

\begin{problem}
Design a 4-bit number subtractor  with carry input and output.

$\diamond$
\end{problem}

\begin{problem}
Design a 4-bit number adder/subtractor with carry input and output. The input {\tt op} designates addition for {\tt op = 0}, and subtract for {\tt op = 1}.

$\diamond$
\end{problem}

\begin{problem}
Fill in the missing lines in Table \ref{sevenSeg} and complete the project partially defined by {\tt sevenSegDys.sv}. Test the correct operation.

$\diamond$
\end{problem}

\begin{problem}
Write a full tester for {\tt dcd.sv} amd run it.

$\diamond$
\end{problem}

\begin{problem}\label{dcdP}
Design in System Verilog a 4-input decoder using the recursive definition provided in Figure \ref{fig8}. Test its correct functioning.

$\diamond$
\end{problem}

\begin{problem}
Design in System Verilog a 4-input demultiplexor using DCD4 designed as solution for Problem \ref{dcdP}. Test its correct functioning.

$\diamond$
\end{problem}

\begin{problem}
Design in System Verilog a 4-input multiplexor using DCD4 designed as solution for Problem \ref{dcdP}. Test its correct functioning

$\diamond$
\end{problem}

\begin{problem}\label{eq}
Consider a decoder with 4 inputs, denoted {\tt A[3:0]}, whose outputs are connected to the selected inputs of a multiplexer with output {\tt F} and with 4 selection inputs, denoted {\tt B[3:0]}. What is the transfer function, {\em F = f(A,B)}, of the system with two 4-bit inputs, {\tt A} and {\tt B} and a one-bit output {\tt F }?

$\diamond$
\end{problem}

\begin{problem}
Provide an optimal solution for a circuit which receives 2 8-bit words, {\tt A[7:0]} and {\tt B[7:0]},  and provide one bit output, EQ, telling if the two words are identical or not:
\begin{equation}
  EQ =
    \begin{cases}
      1 & \text{if  A == B}\\
      0 & \text{if  A !== B}
    \end{cases}
\end{equation}

$\diamond$
\end{problem}


\begin{problem}
Test the behavior of the ALU defined by {\tt alu.sv} in Section \ref{alu1}.

$\diamond$
\end{problem}


\begin{problem}
Modify the module {\tt alu.sv} from Section \ref{alu1} to save the bit {\tt left[0]}, when the right shift is performed, to the output crOut.

$\diamond$
\end{problem}


\begin{problem}\label{upgradeALU}
Add to the module {\tt alu.sv} from Section \ref{alu1} the following functionality and test it:
\begin{itemize}
    \item right shift with {\tt crIn} on the most significant position
    \item arithmetic (right) shift which maintains the sign of the signed integer
    \item rotate left
    \item rotate right
    \item equality comparison: {\tt left == right}
    \item inequality comparison: {\tt left >= right}
    \item select the maximum: {\tt out = (left =< right) ? left : right}
    \item select minimum: {\tt out = (left > right) ? left : right}
\end{itemize}

$\diamond$
\end{problem}


\begin{problem}
Design structurally at the gate level in System Verilog the slice of the simple ALU defined in Figure \ref{fig7}.

$\diamond$
\end{problem}

