\chapter{Performance Optimized Digital Logic}\label{chapt:perf_digital_logic}
\localtoc

\subsection{Comparison of \texttt{when} and \texttt{MuxCase} in Chisel}

The choice between \texttt{when} and \texttt{MuxCase} in Chisel depends on the specific design scenario and synthesis goals. Below is a detailed comparison of these two constructs:

\paragraph{Behavioral and Structural Implications}
\texttt{when} is more natural and intuitive for conditions that resemble software-like \texttt{if-else} logic. It typically generates cascaded conditional logic (e.g., \texttt{if-else} structures) in the hardware, where each condition is checked sequentially. On the other hand, \texttt{MuxCase} represents a priority-free selection logic, akin to a multiplexer, where all conditions are evaluated simultaneously. This structure often synthesizes directly into a multiplexer tree, which can result in better performance and reduced latency for certain designs.

\paragraph{Synthesis Implications}
\textbf{Timing and Performance:}  
For timing, \texttt{when} introduces propagation delays due to the cascading nature of sequential conditions. In contrast, \texttt{MuxCase} generates parallel logic, potentially yielding better timing performance as conditions are evaluated independently.

\textbf{Area:}  
The \texttt{when} construct can produce more compact hardware for sequential conditions since unused conditions are pruned early. However, \texttt{MuxCase} may generate larger hardware if many mutually exclusive conditions are evaluated together because all conditions are checked in parallel.

\textbf{Readability:}  
\texttt{when} is preferable for sequential, human-readable conditions with natural priority, while \texttt{MuxCase} is better suited for a concise representation of mutually exclusive cases.

\paragraph{Specific Considerations}
\textbf{Mutually Exclusive Conditions:}  
If conditions are naturally exclusive, \texttt{MuxCase} is often a better fit. For example:
\[
\texttt{val result = MuxCase(0.U, Seq(}
\]
\[
\quad (\texttt{opcode === ADD} \rightarrow \texttt{(a + b),}
\]
\[
\quad (\texttt{opcode === SUB} \rightarrow \texttt{(a - b))}
\]
\texttt{))}
This translates to a 3-to-1 multiplexer in hardware.

\textbf{Priority Logic:}  
When conditions have a natural priority (e.g., an \texttt{else} case that catches all unhandled conditions), \texttt{when} is preferable. For example:
\[
\texttt{when(opcode === ADD) \{}
\]
\[
\quad \texttt{result := a + b}
\]
\[
\texttt{\}.elsewhen(opcode === SUB) \{}
\]
\[
\quad \texttt{result := a - b}
\]
\[
\texttt{\}.otherwise \{}
\]
\[
\quad \texttt{result := 0.U}
\]
\texttt{\}}  
This logic represents a chain of conditional checks in hardware.

\paragraph{Recommendation}
\texttt{when} should be used for conditions with priority, sequential evaluation, or better readability when handling control logic. In contrast, \texttt{MuxCase} is ideal for selection logic like multiplexer trees, where all conditions are naturally parallel and priority-free. By carefully analyzing the nature of your conditions and their synthesis implications, you can choose the construct that aligns best with your performance and area optimization goals.


\subsubsection{Switch Synthesis in Chisel}
In Chisel, a \texttt{switch} statement is a construct used for multi-way branching based on a condition or value. The hardware synthesized from a \texttt{switch} statement depends on the conditions and the synthesis toolchain. Generally, it can result in one of the following hardware structures:

A \textbf{multiplexer tree} is generated if the \texttt{switch} conditions are mutually exclusive, with each case corresponding to a distinct value of the selector. For instance:
\[
\texttt{
val result = Wire(UInt(8.W)) \\
switch(selector) \{ \\
\quad is("b00".U) \{ result := a \} \\
\quad is("b01".U) \{ result := b \} \\
\quad is("b10".U) \{ result := c \} \\
\quad is("b11".U) \{ result := d \} \\
\}
}
\]
This would synthesize into a 4-to-1 multiplexer controlled by the value of \texttt{selector}.

When the \texttt{switch} includes overlapping or non-mutually-exclusive conditions, \textbf{priority logic} is typically generated. Priority logic evaluates cases sequentially, giving precedence to earlier conditions. For wide selectors or more complex logic, synthesis tools may generate \textbf{decoder logic} to translate the selector into one-hot or prioritized enable signals that control the corresponding cases.

\subsubsection{Comparison with \texttt{MuxCase} and \texttt{when}}
\texttt{MuxCase} is more explicit and directly creates a multiplexer, making it compact and efficient for mutually exclusive conditions. \texttt{when} constructs explicitly generate priority logic. In contrast, \texttt{switch} can represent either multiplexers or priority logic depending on the use case, providing a flexible but less predictable synthesis.

\subsubsection{Limitations of \texttt{switch}}
While \texttt{switch} provides a clean and readable way to handle multi-way branching, it is not always recommended for large or complex selectors. For wide selectors, the generated decoder logic can result in significant area and timing overhead. Additionally, overlapping or non-mutually-exclusive cases can inadvertently lead to priority chains that increase latency and complicate timing analysis. Designers must carefully evaluate the trade-offs between simplicity in the design code and hardware efficiency, especially for performance-critical applications.


%**********************************
\section{Carry Lookahead Adder}
%*********************************
Carry Look-Ahead Adders (CLAs) are designed to overcome the performance bottlenecks of ripple-carry adders by reducing the dependency on sequential carry propagation. Instead of waiting for each bit's carry to be computed, the CLA computes carries in parallel using the generate (\texttt{G}) and propagate (\texttt{P}) signals for each bit. These signals allow the adder to determine whether a carry will be generated or propagated at each stage. Using these signals, the carry for each bit position can be calculated directly through a set of Boolean equations, significantly reducing the delay to logarithmic time relative to the adder's width. This design is particularly effective for high-speed arithmetic operations in processors and digital signal processing, where minimizing latency is important. However, the parallel computation and additional logic circuitry result in increased hardware complexity compared to simpler adders.

%******************************
\section{Carry Select Adder}
%******************************
Carry Select Adders (CSAs) improve upon the delay of ripple-carry adders by precomputing the sum and carry outputs for both possible carry-in values (0 and 1) in parallel. Each stage of the adder computes two possible results based on these assumptions, and a multiplexer is used to select the correct result once the carry-in is determined. This approach significantly reduces the propagation delay associated with sequential carry computation while maintaining a relatively straightforward design. Although CSAs require additional hardware resources for the parallel computations and multiplexers, they achieve a good trade-off between speed and area, making them well-suited for medium-sized arithmetic units where both latency and complexity are important considerations.

%*****************************
\section{Carry Save Adder}
%*****************************
Carry-save adders are specialized adders used primarily in multiplication. Unlike conventional adders, which propagate carries through the stages immediately, CSAs represent the result as two separate numbers: a partial sum and a carry vector. These two outputs are then combined in a subsequent addition step, often using a more efficient adder such as a carry-lookahead adder. By deferring the carry propagation, CSAs significantly reduce the delay for intermediate steps, making them highly efficient for scenarios where multiple additions must be performed simultaneously. Examples include the Wallace tree \cite{wallace1964} and Dadda tree \cite{dadda1965}. This efficiency is achieved at the cost of additional logic to manage the separate sum and carry outputs.






%*****************************
\section{Kogge-Stone Adder} \label{sec:kogge-stone-adder}
%*****************************
Parallel prefix adders are a class of high-performance adders designed to optimize the computation of carry signals in binary addition, significantly reducing delay compared to simpler adder designs like ripple-carry adders. They achieve this by dividing the addition process into three distinct stages: generating initial propagate and generate signals for each bit, computing intermediate carries through a tree-like prefix structure, and finally calculating the sum based on these carry signals. The tree structure allows the carry signals to be computed in parallel, with a logarithmic time complexity relative to the number of bits, making these adders highly efficient for large bit-width operations. These adders are widely used in modern high-speed computing systems and processors.

The Kogge-Stone Adder is a parallel prefix adder known for its high performance and suitability for large bit-width arithmetic operations. It minimizes the propagation delay using a tree-like structure to compute carry signals in a logarithmic number of stages relative to the bit width. The design uses generate and propagate signals to recursively compute carries in parallel, ensuring that the carry for any given bit is available after just $\log_2(n)$ stages, where $n$ is the bit width. While this structure significantly enhances speed, it requires substantial wiring and additional resources, making it resource-intensive compared to simpler designs like ripple-carry adders. Due to its efficiency in reducing delay, the Kogge-Stone Adder is frequently employed in high-performance processors and digital systems requiring fast arithmetic computations  \cite{kogge1973parallel}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{{\bf {\em Multipliers}}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%