%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%
Chisel (\textbf{C}onstructing \textbf{H}ardware \textbf{I}n a \textbf{S}cala \textbf{E}mbedded \textbf{L}anguage) is a hardware design language embedded within Scala. It allows hardware designers to write complex, parameterizable, and reusable hardware components using high-level programming constructs \cite{bachrach2012chisel}. 

This Appendix is a very brief introduction to Chisel primarily for designers to reference features quickly. For processor design, \citep{chisel:book} is a highly recommended free book. The Chisel Cheat Sheet is also helpful \cite{chiselcheatsheet}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Key Features of Chisel}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Chisel is a domain-specific language (DSL) for hardware design, using Scala's object-oriented and functional programming features \cite{chisel-doc}. It compiles to synthesizable Verilog, integrating with traditional hardware development flows.

\paragraph{Hardware Construction Language.} High-level abstractions for hardware components.
\paragraph{Parameterization.} Easy creation of generic and reusable modules.
\paragraph{Strong Typing.} Reduces errors through rigorous type checking.
\paragraph{Functional Programming.} Utilizes Scala's functional paradigms for concise code.


%%%%%%%%%%%%%%%%%
\section{Syntax}
%%%%%%%%%%%%%%%%

%************************
\subsection{Scala Basics}
%************************
Understanding basic Scala syntax is needed for Chisel programming \cite{odersky2008programming}.

\paragraph{Immutable Variables (\texttt{val}).} Cannot be reassigned.
\paragraph{Mutable Variables (\texttt{var}).} Can be reassigned.
\paragraph{Functions (\texttt{def}).} Defined using the \texttt{def} keyword.
\paragraph{Classes and Objects.} Used to define hardware modules and singletons.

%******************************
\subsection{Import Statements}
%*****************************
Include necessary Chisel libraries at the beginning of your file:

\begin{chiselcode}
import chisel3._
import chisel3.util._    
\end{chiselcode}

%%%%%%%%%%%%%%%%%%%%
\section{Data Types}
%%%%%%%%%%%%%%%%%%%%
Chisel provides specialized data types for hardware representation \cite{chisel-doc}.

%***********************
\subsection{Base Types}
%**********************
\paragraph{Bool.} Single-bit boolean (\texttt{true}/\texttt{false}). Declaration: \texttt{val myBool = Bool()}

\paragraph{UInt.} Unsigned integer with specified width. Declaration: \texttt{val myUInt = UInt(8.W)}

\paragraph{SInt.} Signed integer with specified width. Declaration: \texttt{val mySInt = SInt(16.W)}

\paragraph{FixedPoint.} Fixed-point number with total and fractional bits. Declaration: \texttt{val myFixed = FixedPoint(16.W, 8.BP)}

\paragraph{Clock.} Represents a clock signal.

\paragraph{Reset.} Represents a reset signal.

%****************************
\subsection{Aggregate Types}
%***************************
\paragraph{Bundle.} Custom data structures combining multiple fields. Declaration: \texttt{class MyBundle extends Bundle \{ val a = UInt(8.W) \}}

\paragraph{Vec.} Indexable collection of elements of the same type. Declaration: \texttt{val myVec = Vec(4, UInt(8.W))}

%********************
\subsection{Literals} 
%********************
\paragraph{Unsigned Integer.}  \texttt{10.U}, \texttt{3.U(8.W)}

\paragraph{Signed Integer.} \texttt{-5.S}, \texttt{-2.S(8.W)}

\paragraph{Boolean.} \texttt{true.B}, \texttt{false.B}

%*********************
\subsection{Operators} 
%*********************
\paragraph{Arithmetic (\texttt{+, -, *, /, \%}).} \texttt{+} (addition), \texttt{-} (subtraction), \texttt{*} (multiplication), \texttt{/} (division), \texttt{\%} (modulo operation)

\paragraph{Logical (\texttt{\&, |, \^, !}).} \texttt{\&} (logical AND), \texttt{|} (logical OR), \texttt{\^{}} (logical XOR), \texttt{!} (logical NOT)

\paragraph{Comparison (\texttt{===, =/=, <, <=, >, >=}).} \texttt{===} (equal to), \texttt{=/=} (not equal to), \texttt{<} (less than), \texttt{<=} (less than or equal to), \texttt{>} (greater than), \texttt{>=} (greater than or equal to)

\paragraph{Bitwise Reduction (\texttt{\&, |, \^}).} \texttt{\&} (bitwise AND reduction), \texttt{|} (bitwise OR reduction), \texttt{\^{}} (bitwise XOR reduction). These operators are applied to a single operand.

\paragraph{Shift (\texttt{<<, >>}).} \texttt{<<} (left shift), \texttt{>>} (right shift)

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Modules and Hierarchy}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%****************************
\subsection{Defining Modules}
%****************************
Modules are defined by extending the \texttt{Module} class \cite{chisel-doc}.

\begin{chiselcode}
class MyModule extends Module {
  // Module implementation
}    
\end{chiselcode}

%******************************
\subsection{Input/Output Ports}
%*******************************
Ports are declared using the \texttt{IO} method with a \texttt{Bundle}.
\begin{chiselcode}
class MyModule extends Module {
  val io = IO(new Bundle {
    val in  = Input(UInt(8.W))
    val out = Output(UInt(8.W))
  })
  // Module logic
}    
\end{chiselcode}

%*********************************
\subsection{Instantiating Modules}
%*********************************
Modules can be instantiated within other modules.
\begin{chiselcode}
val subModule = Module(new SubModule)
subModule.io.in := io.in
io.out := subModule.io.out
\end{chiselcode}

%*********************************
\subsection{Parameterized Modules}
%*********************************
Modules can be parameterized for reusability.
\begin{chiselcode}
class MyModule(val width: Int) extends Module {
  val io = IO(new Bundle {
    val in  = Input(UInt(width.W))
    val out = Output(UInt(width.W))
  })
  // Module logic
}    
\end{chiselcode}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Wires and Registers}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Wires represent combinational logic signals \cite{chisel-doc}.
\begin{chiselcode}
val tempWire = Wire(UInt(8.W))
tempWire := io.in + 1.U    
\end{chiselcode}

Registers store state across clock cycles.
\begin{itemize}
    \item \texttt{Reg}: Creates a register without an initial value.
        \newline Declaration: \texttt{val myReg = Reg(UInt(8.W))}
    \item \texttt{RegInit}: Creates a register with an initial value.
        \newline Declaration: \texttt{val myReg = RegInit(0.U(8.W))}
\end{itemize}

Registers are updated using non-blocking assignment.
\begin{chiselcode}
myReg := myReg + 1.U    
\end{chiselcode}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Control Structures}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%************************************
\subsection{\texttt{when} Statements}
%************************************
Conditional hardware execution.
\begin{chiselcode}
when (condition) {
  // Actions when condition is true
} .elsewhen (anotherCondition) {
  // Actions when anotherCondition is true
} .otherwise {
  // Actions when all conditions are false
}
\end{chiselcode}

%**************************************
\subsection{\texttt{switch} Statements}
%**************************************
Multi-way conditional execution.
\begin{chiselcode}
switch(selector) {
  is(value1) {
    // Actions for value1
  }
  is(value2) {
    // Actions for value2
  }
}
\end{chiselcode}

%*********************************
\subsection{\texttt{Mux} Function}
%*********************************
Multiplexer for conditional selection.
\begin{chiselcode}
val result = Mux(condition, trueValue, falseValue)
\end{chiselcode}

%*******************
\subsection{Loops}
%******************
Used for hardware generation, not for creating sequential logic \cite{chisel-tutorial}.
\begin{chiselcode}
for (i <- 0 until n) {
  // Generate n instances of hardware
}
\end{chiselcode}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Functions and Procedures}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%******************************
\subsection{Defining Functions}
%******************************
Functions encapsulate reusable logic.
\begin{chiselcode}
def add(a: UInt, b: UInt): UInt = {
  a + b
}
\end{chiselcode}

%*******************************
\subsection{Invoking Functions}
%******************************
\begin{chiselcode}
val sum = add(io.in1, io.in2)
\end{chiselcode}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Parameterized Types}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Types can be parameterized for flexibility.
\begin{chiselcode}
def myFunction[T <: Data](in: T): T = {
  // Function body
}
\end{chiselcode}

The function \texttt{myFunction} is a generic method in Chisel that operates on a parameterized data type \texttt{T}. The parameter \texttt{T} is constrained to extend the \texttt{Data} type, which is the base class for all hardware types in Chisel, such as \texttt{UInt}, \texttt{SInt}, or \texttt{Bundle}. This ensures that \texttt{myFunction} can accept any valid Chisel hardware type as its input. The method takes a single input argument \texttt{in} of type \texttt{T} and returns a value of the same type \texttt{T}. While the function body is not specified, this structure is commonly used to define reusable, type-safe operations on hardware constructs, enabling modular and flexible hardware design.

%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Generic Modules}
%%%%%%%%%%%%%%%%%%%%%%%%%%
Modules can be generic over data types.
\begin{chiselcode}
class MyGenericModule[T <: Data](genType: T) extends Module {
  val io = IO(new Bundle {
    val in  = Input(genType)
    val out = Output(genType)
  })
  // Module logic
}
\end{chiselcode}

The \texttt{MyGenericModule} class is a parameterized hardware module in Chisel. It extends the \texttt{Module} base class and takes a type parameter \texttt{T}, constrained to be a subtype of \texttt{Data}, which ensures that it can represent any Chisel hardware type such as \texttt{UInt}, \texttt{SInt}, or custom \texttt{Bundle}s. The parameter \texttt{genType} is passed to the constructor and defines the specific type of data the module will operate on. Inside the module, the \texttt{io} interface is declared using a \texttt{Bundle}, containing an input port \texttt{in} and an output port \texttt{out}, both of which use the generic type \texttt{genType}. This design allows the module to be highly reusable and adaptable, as it can handle different data types depending on the parameter provided during instantiation. The comment \texttt{// Module logic} indicates where the functional implementation of the module would be added.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Hardware Construction in Chisel}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%***********************************
\subsection{Vectors (\texttt{Vec})}
%***********************************
Indexable collections of elements.
\begin{chiselcode}
val myVec = VecInit(Seq.fill(4)(0.U(8.W)))
myVec(0) := io.in
\end{chiselcode}

%************************************
\subsection{Memories (\texttt{Mem})}
%***********************************
Memory elements for storage.
\begin{chiselcode}
val myMem = Mem(1024, UInt(8.W))
val dataOut = myMem.read(addr)
myMem.write(addr, dataIn)
\end{chiselcode}

%*******************
\subsection{Bundles}
%*******************
Custom data structures grouping multiple fields.
\begin{chiselcode}
class MyBundle extends Bundle {
  val field1 = UInt(8.W)
  val field2 = Bool()
}
\end{chiselcode}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Testing and Verification}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{ChiselTest Framework}
%********************************
ChiselTest provides tools for testing Chisel modules \cite{chiseltest}.
\begin{chiselcode}
import chiseltest._
import org.scalatest.flatspec.AnyFlatSpec

class MyModuleTest extends AnyFlatSpec with ChiselScalatestTester {
  "MyModule" should "perform correctly" in {
    test(new MyModule) { c =>
      c.io.in.poke(0.U)
      c.clock.step(1)
      c.io.out.expect(0.U)
    }
  }
}
\end{chiselcode}

%**********************
\subsection{Assertions}
%**********************
In-module checks using \texttt{assert}.
\begin{chiselcode}
assert(io.out === expectedValue)
\end{chiselcode}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Peculiarities of Chisel}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%**************************************
\subsection{Chisel is Embedded in Scala}
%**************************************
Chisel code is Scala code. This means:

\begin{itemize}
    \item Full access to Scala's features.
    \item Potential confusion between hardware constructs and software constructs.
\end{itemize}

%***********************************
\subsection{Delayed Execution Model}
%***********************************
Chisel builds a hardware graph during elaboration, which may lead to unexpected behavior if side effects are not managed properly \cite{chisel-doc}.

%********************************
\subsection{Assignment Operators}
%********************************
\begin{itemize}
    \item \texttt{:=} Used for updating hardware signals.
    \item \texttt{=} Used for variable assignment in Scala.
\end{itemize}

%***********************
\subsection{Type System}
%***********************
Chisel enforces strict type checking, which may differ from SystemVerilog or VHDL.

%******************************************************
\subsection{Combining Hardware and Software Constructs}
%******************************************************
Be cautious when mixing Scala control structures (e.g., \texttt{if}, \texttt{for}) with Chisel constructs (\texttt{when}, \texttt{switch}) \cite{chisel-tutorial}.

%*****************************************
\subsection{Hardware vs. Software Mindset}
%*****************************************
Designers must think in terms of hardware concurrency, even when writing code that resembles sequential software.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Installing Chisel on Ubuntu}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{enumerate}
    \item \textbf{Install Java Development Kit (JDK):} \\
    Chisel is built on Scala, which requires the Java Development Kit (JDK). You have two options to install and manage the JDK:
    \begin{enumerate}
        \item \textbf{Option 1: Install the Default JDK:} \\
        Run the following commands to install the default JDK package:
        \begin{bashcode}
sudo apt update && sudo apt install default-jdk
        \end{bashcode}

        \item \textbf{Option 2: Use Maven to Manage JDK Versions:} \\
        If you already use Maven, ensure you have a JDK installed and configured in your Maven settings. You can verify your JDK version with:
        \begin{bashcode}
mvn --version
        \end{bashcode}
        If a JDK is not configured, follow Maven's documentation to set up a compatible JDK.
    \end{enumerate}

    \item \textbf{Install Scala Build Tool (sbt):} \\
    Chisel projects are typically managed using sbt. To install sbt, execute:
    \begin{bashcode}
echo "deb https://repo.scala-sbt.org/scalasbt/debian all main" | \
sudo tee /etc/apt/sources.list.d/sbt.list && \
curl -sL "https://keyserver.ubuntu.com/pks/lookup? \ 
op=get&search=0x99E82A75642AC823" | \
sudo apt-key add && \
sudo apt update && \
sudo apt install sbt
    \end{bashcode}

    \item \textbf{Clone this book:} \\
    Or use the template at 
    \begin{bashcode}
https://github.com/chipsalliance/chisel
    \end{bashcode}

    \item \textbf{Build and Run Your Chisel Project:} \\
    Navigate to your project's directory and use sbt to compile and run it:
    \begin{bashcode}
cd your-chisel-project && sbt test
    \end{bashcode}
\end{enumerate}
