\documentclass[11pt]{book}

\makeatletter
\let\cleardoublepage\clearpage
\makeatother

\usepackage{lipsum}
\usepackage{comment}
\usepackage{verbatim}
\usepackage{latexcad}  % Position sensitive
\usepackage[maxfloats=25]{morefloats}
\usepackage[english]{babel}

\usepackage{graphicx}
\graphicspath{{}{Images/}}  % Look in ./ and Images/
\usepackage{xcolor}  % Make cells of an array a specific color

\usepackage{makeidx}
\usepackage{mathptmx}



\usepackage{amsfonts}
\usepackage[hyphens]{url}
\usepackage[utf8]{inputenc}

\usepackage{fontspec}
\setmainfont{Times New Roman}
\usepackage{fontspec} % necesar pentru Unicode
%\setmainfont{Noto Sans CJK SC} 
%\usepackage{cmbright}
%\usepackage{lmodern}
%\usepackage{epic,eepic}
%\usepackage{pstricks}

\usepackage{mathptmx, framed, xcolor}

\definecolor{shadecolor}{rgb}{0.97,0.97,0.97} % light gray code background
%\definecolor{shadecolor}{rgb}{1.0,1.0,1.0} % white code background (saves ink)

\usepackage{amsmath}
\usepackage{float}
\usepackage{array}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{booktabs}   %\toprule
\usepackage{makecell}

\usepackage{circuitikz} %circuit diagrams
\usepackage{tikz}

\usepackage{fancyhdr}

\usepackage{longtable}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Table of Contents Formatting
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{listings}
\usepackage{shorttoc}   % Creates a short TOC
\usepackage{tocbibind}  % Adds Bibliography/Index to TOC
\usepackage{tocloft}    % Custom TOC spacing
\usepackage{etoc}       % Enhanced TOC management
\usepackage{etoolbox}   % Utility package for patching commands

% Register List of Listings with tocloft
\newcommand{\listoflistingsheading}{\section*{\lstlistlistingname}} % Optional heading

% Define lengths for List of Listings title spacing
\newlength{\cftbeforeloltitleskip}
\newlength{\cftafterloltitleskip}
\newlength{\cftchapterskip}

\setlength{\cftbeforeloltitleskip}{1em}  % Space before title
\setlength{\cftafterloltitleskip}{0.5em} % Space after title
\setlength{\cftbeforepartskip}{2em} % Space before part entries

% Patch \lstlistoflistings for spacing
\makeatletter
\patchcmd{\lstlistoflistings}
  {\@starttoc{lol}}
  {\vspace{\cftbeforeloltitleskip}\@starttoc{lol}\vspace{\cftafterloltitleskip}}
  {}{}
\makeatother

\newcommand{\localtoc}{%
    \begingroup
    \setcounter{tocdepth}{2} % Set depth to include sections and subsections
    \localtableofcontents    % Call the local TOC
    \newpage                 % Add a page break
    \setcounter{tocdepth}{5} % Set depth to include sections and subsections
    \endgroup
}



%%%%%%%%%%%%%%%%%%%%%%
% Citations
%%%%%%%%%%%%%%%%%%%%%
% Swap the behavior of \cite and \citep
% \cite{} -> (Author, 2024)
 % \citep{}  -> Author (2024)
\usepackage[authoryear, round]{natbib} 
\let\citeold\cite
\let\cite\citep
\let\citep\citeold


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Code Formatting
\input{00_code_formatting}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\setlength{\textheight}{8.25in} \setlength{\textwidth}{6in}
\topmargin 323mm \advance \topmargin -\textheight \divide
\topmargin by 4 \advance \topmargin -1.0in \leftmargin 216mm
\advance \leftmargin -\textwidth \divide \leftmargin by 2 \advance
\leftmargin -1.0in \oddsidemargin \leftmargin \evensidemargin
\leftmargin


%******************
% List Parameters
%*****************
%\usepackage{etoolbox} % For patching commands
\usepackage{enumitem} % Required for customizing list environments
\setlist[itemize,1]{ 
    %label=\scalebox{1.5}{\textbullet},
    left=3em,           % Increase indent 
    topsep=0.25em,       % Space above the list
    partopsep=0em,      % Additional space above if starting a new paragraph
    parsep=0.5em,         % Space between paragraphs inside the list
    itemsep=0.5em,       % Space between list items
    listparindent=0em,    % Space for paragraphs inside the list
    after=\vspace{0.5em}\noindent % Remove indentation after all lists
}
\setlist[itemize,2]{
    label=$\diamond$,
    left=3em,           % Increase indent 
    topsep=0.25em,       % Space above the list
    partopsep=0em,      % Additional space above if starting a new paragraph
    parsep=0.5em,         % Space between paragraphs inside the list
    itemsep=0.5em,       % Space between list items
    listparindent=0em,    % Space for paragraphs inside the list
    %after=\vspace{1em}\noindent % Remove indentation after all lists
}

\setlist[enumerate,1]{ % Top-level list
    label=\arabic*),   % 1), 2), 3), ...
    left=3em,         
    topsep=0.25em,     
    partopsep=0em,     
    parsep=0.5em,        
    itemsep=0.5em,     
    listparindent=0em, 
    after=\vspace{0.5em}\noindent
}
\setlist[enumerate,2]{ % Second-level nested list
    label=\alph*),     % a), b), c), ...
    topsep=0.25em,     
    partopsep=0em,     
    parsep=0.5em,        
    itemsep=0.5em,     
    listparindent=0em, 
    %after=\vspace{1em}\noindent
}

%%%%%%%%%%%
\makeindex
%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%
% DOCUMENT
%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\pagestyle{fancy}
\sloppy

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% HEADER / FOOTER
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\fancyhead{} % Clear all header fields
%\fancyhead[RO,LE]{\footnotesize \textbf{Gheorghe M. \c{S}tefan: \emph{Structured Computer Architecture}}} % Right on odd pages, left on even pages
\fancyhead[RO,LE]{\footnotesize \textbf{\emph{Structured Computer Architecture}}} % Right on odd pages, left on even pages
\fancyhead[LO]{\scriptsize \leftmark} % Chapter title in the left header on odd pages
\fancyhead[RE]{\scriptsize \leftmark} % Chapter title in the right header on even pages
\fancyfoot{} % Clear all footer fields
\fancyfoot[RO]{\footnotesize \thepage} % Right for odd pages
\fancyfoot[LE]{\footnotesize \thepage} % Left for even pages


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Title
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{{\Huge {\bf Structured Computer Architecture\\
An Order-Based Approach}}
}
\author{{\LARGE {\bf \em John Glossner and Gheorghe M. \c{S}tefan}}}
\date{
%\date{\today \\ \vspace{1em} \vspace{1em} 
\placedrawing*[h]{figures/YYY.LP}
{\em First Edition}
}



\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Front matter
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\pagenumbering{roman}
%\part*{FRONT MATTER}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Copyright
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter*{Copyright}
\addcontentsline{toc}{chapter}{Copyright}

\section*{Copyright Notice}

\vspace{\baselineskip}\par

\addcontentsline{toc}{section}{All text content is Copyright © 2025-2026 CC-BY-NC-ND 4.0 by John Glossner and Gheorghe M. \c{S}tefan until our deaths when it shall become CC-BY-NC-SA 4.0. }

All text content is Copyright © 2025-2026 by John Glossner and Gheorghe M. \c{S}tefan licensed under a Creative Commons Attribution-NonCommercial-NoDerivatives 4.0 International License (CC BY-NC-ND 4.0). \url{https://creativecommons.org/licenses/by-nc-nd/4.0/deed.en}

\vspace{\baselineskip}\par
\noindent Upon our deaths, the text copyright shall convert to © 2025-2026 by John Glossner and Gheorghe M. \c{S}tefan licensed under a Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License (CC BY-NC-SA 4.0). \url{https://creativecommons.org/licenses/by-nc-sa/4.0/deed.en}

\vspace{\baselineskip}\par
\noindent Some portions of this document contain edited text originally generated by ChatGPT. 

\section*{Code License}
\addcontentsline{toc}{section}{Chisel Scala Code License - Solderpad Hardware License v2.1}
Chisel Scala code is licensed under the Solderpad Hardware License v2.1. See https://solderpad.org/licenses/SHL-2.1/ for details.

\vspace{\baselineskip}\par
\addcontentsline{toc}{section}{Hand Written Verilog Code License - CERN-OHL-S v2}
\noindent SystemVerilog and Verilog code are licensed under CERN-OHL-S v2. See https://cern.ch/cern-ohl for details.

\section*{Typesetting}
This document was prepared with \LaTeX $\, 2_{\epsilon}$

\bigskip

\bigskip

ISBN 9798195042912

Imprint: Independently published

\bigskip

Cover design by: Elisabeth Jacob

Artwork generated using AI, inspired by the abstract style of Paul Klee.


\sloppy

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Dedication
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter*{Dedication}

\vspace{\baselineskip}\par \noindent
John: {\it To my beloved wife Lisa who is and has been my life partner since high school.}

\vspace{\baselineskip}\par \noindent
George: {\it To Mihaela.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Epigraph
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\chapter*{The Halting Problem}

\chapter*{Motivation}
\addcontentsline{toc}{chapter}{Motivation}

\vspace{\baselineskip}\par \noindent
\begin{quote}
{\bf Question}: {\em Can we have a procedure that, receiving a program and the data it processes, decides if the program will stop executing in a finite time?}
\end{quote}

\vspace{\baselineskip}\par \noindent
This (halting) problem was formulated in computational terms in 1936 by Alonzo Church, Stephen Kleene, Emil Post, and Alan Turing as a reaction to the incompleteness theorem published by Kurt G\" odel in 1931. 

\begin{quote}
{\bf Answer: {\em No!}}
\end{quote}

\vspace{\baselineskip}\par \noindent
And thus, the most important negative result in the history of mathematics underpins the science of computation.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Forward
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Placeholder if we find someone to write one
%\chapter*{Forward}
%\addcontentsline{toc}{chapter}{Forward}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Preface
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

This textbook arose from an introductory course on Computer Architecture taught at Rivier University in the Fall of 2024. %It draws on Gheorghe M. \c{S}tefan's years of teaching Computer Architecture at the University of Bucharest.

Unlike other Computer Architecture textbooks, this book approaches the subject from a Computer Science theoretical perspective. Using an ordered approach, we start with the most basic 0-order computing system - one with stateless operations. We then describe computing systems in terms of their feedback loop orders progressing order-by-order \cite{stefan24}. Table \ref{tab:orders_of_computing_systems} shows the  progression of computing systems feedback loops.

%***********************************
% TABLE: Order of Computing Systems
%***********************************
\begin{table}[h!]
\centering
\renewcommand{\arraystretch}{1.2} % Increase row height for vertical centering
\begin{tabular}{|>{\centering\arraybackslash}m{1.8cm} | >{\centering\arraybackslash}m{4.2cm} | >{\centering\arraybackslash}m{3.7cm} | >{\centering\arraybackslash}m{3.8cm}|}
\hline
\textbf{Order} & \textbf{Definition} & \textbf{Components} & \textbf{Example} \\ \hline

0-Order & No-loop combinational circuits. & Logic gates (AND, OR, NOT, ROM). & Multiplexer, adder, ALU. \\ \hline

1st-Order & One-loop memory circuits & Latches, FFs, registers, RAM, SRAM. & D flip-flop, Register file \\ \hline

2nd-Order & Two-loop automata circuits. & Mealy, Moore automaton & FSMs, Counters,  ALUs with register files. \\ \hline

3rd-Order & Three-loop processing units. & ALU, control units, registers. & Elementary processor, Simple microprocessor. \\ \hline

4th-Order & Four-loop computer with one memory & Program and Data in single memory. & von Neumann abstract model computer. \\ \hline

5th-Order & Five-loop computer with two memories. & Separate Program and Data memories. & Harvard abstract model computer. \\ \hline

nth-Order  & N-loop, no global loop many-cell systems. & Controller, Map Array & Map Parallel Accelerator  \\ \hline

1st  Global Loop & N-loop \&  1st global loop many-cell systems. & Controller, Map Array, Reduce loop & MapReduce Parallel Accelerator  \\ \hline

2nd  Global Loop & N-loop \& 2nd global loop many-cell systems. & Controller, Map Array, Reduce loop, Scan loop & MapScanReduce Parallel Accelerator \\ \hline

\end{tabular}
\caption{Orders of Digital Systems from Gates to Parallel Accelerators.}
\label{tab:orders_of_computing_systems}
\end{table}

Part I of this book provides a mathematical foundation for the number systems used in computing systems. Part II covers orders 0 through 3 - basic combinational logic through a processor. Part III covers the theory and implementation of computing, including parallel heterogeneous systems. Part IV refers to practical aspects related to the design of frequently used combinational circuits, the design of processors, and the use of heterogeneous computing systems. 

This book also differs from one that deals with digital systems in that it pays attention only to those digital systems that contribute directly to the simplest structures that allow the enunciation of general principles related to the architecture of computing systems. A minimal path has been identified through the exceptionally large set of digital structures that leads to the generic computing structures that allowed us to achieve the proposed target.

In our opinion, an advantage of the order-based approach is that it roughly follows the historical and structural development of computing systems. We also ground these developments in the theoretical models that provided for their development.

Although we write primarily for our students, we have both industry experience and have shipped multiple machines in volume. The code we provide bridges the gap between theoretical models and commercial implementations. The code starts very simple using behavioral constructs but progresses towards code that can be used in commercial machines. Over time, it is our hope that these examples continue to expand.

The descriptions of the structures and the simulation of their behavior are written in both SystemVerilog and Chisel HDL. The Chisel code has unit tests and has been tested using Chisel 6.6.0. Some of the SystemVerilog code has unit tests. All SystemVerilog codes have been validated using the AMD/Xilinx Vivado Design Suite \footnote{\url{https://www.xilinx.com/content/xilinx/en/support/download.html/}}. It should also be executed using online tools such as EDA Playground \footnote{\url{https://www.edaplayground.com}}.

Because this book's code is executable, it should not contain any syntax errors. The same files used for design descriptions are directly imported into the book.

This preliminary release we know is quite rough. However, it will become polished with time. The best way to help this effort is to send pull requests to \url{https://github.com/glossner/Structured_Computer_Architecture}. 

\vspace{\baselineskip}\par \noindent
John Glossner and Gheorghe M. \c{S}tefan
\vspace{\baselineskip}\par \noindent
June 3rd, 2026

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Acknowledgments
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\chapter*{Acknowledgments}
%\addcontentsline{toc}{chapter}{Acknowledgments}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Short Table of Contents
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\shorttoc{Contents}{1} % Only sections

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Table of Contents
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand*\contentsname{Contents (detailed)}
\setcounter{secnumdepth}{3}
\setcounter{tocdepth}{4}
\tableofcontents

\listoffigures
%\addcontentsline{toc}{chapter}{List of Figures}

\listoftables
%\addcontentsline{toc}{chapter}{List of Tables}

\lstlistoflistings



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MAIN DOCUMENT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\mainmatter
\pagenumbering{arabic}

\setcounter{tocdepth}{3}% local TOC displays only sections
\setcounter{secnumdepth}{3}
\etocsettocstyle{}{} % No title or end content in local table of contents


\renewcommand*{\indexname}{Glossary}

\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{exemplu}{Example}[chapter]
\newtheorem{definition}{Definition}[chapter]
\newtheorem{problem}{Problem}[chapter]
\newtheorem{project}{Project}[chapter]
\newtheorem{simulation}{VeriSim}[chapter]
\newtheorem{verisum}{SystemVerilogSummary}%[chapter]
\newtheorem{corollary}{Corollary}[chapter]
\newtheorem{proof}{Proof}[chapter]


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\part{NUMBER SYSTEMS AND REPRESENTATIONS}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\include{01_01_number_systems}
\include{01_02_formal_notation}
\include{01_03_unsigned_integers}
\include{01_04_signed_integers}
\include{01_05_floating_point}
\include{01_06_character_encodings}
\include{01_07_part1_summary}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\part{FROM GATES TO PROCESSORS}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\include{02_08_combinational_circuits_0-OS}
\include{02_09_memory_circuits_1-OS}
\include{02_10_automata_2-OS}
\include{02_11_processors_3-OS}

%################
\part{COMPUTING}
%################
\include{03_12_von_neumann_4-OS}
\include{03_13_harvard_5-OS}
\include{03_14_parallel_computing_N-OS}

%######################
\part{PRACTICAL DESIGNS}
%######################
\include{04_15_Performance_Optimized_Digital_Logic}
\include{04_16_ILP}
\include{04_17_0_ToyRISC}
%\include{04_18_RISCV}
%\include{04_19_RISCV_multithreaded}
\include{04_20_pRISC_heterogeneous}


%\include{09_open_problems}

%%%%%%%%%%%%%
% APPENDICES
%%%%%%%%%%%%%
\clearpage
\addcontentsline{toc}{part}{APPENDIXES}
\part*{APPENDIXES}
\appendix

\include{A01_0_Hardware_Description_Language}
\include{A02_physical_gates}
%\include{Axx_simulations}
%\include{Axy_verilogFIFO} %it is proposed as problem in automata chapter

%\include{Latex/A03_Linear_Algebra}

% To Remove Code Listings from the Listings section, uncomment {lol}
%\addtocontents{toc}{\protect\setcounter{tocdepth}{2}}
%\addtocontents{lol}{\protect\setcounter{tocdepth}{-1}}
%\input{A99_Code_Listings}
%\addtocontents{toc}{\protect\setcounter{tocdepth}{4}} % Restore TOC depth
%\addtocontents{lol}{\protect\setcounter{tocdepth}{2}}


%**********************
% Bibliography
%**********************
\clearpage
\bibliographystyle{apalike}
%\bibliographystyle{plainnat}
\bibliography{bibliography} 


\clearpage
\printindex

\end{document}

