\chapter{Computers: Five-Loop, Fifth-Order Systems (5-OS)}\label{lect6a}
\localtoc

We will practice the processor concept in the context provided by the fifth loop, the one that defines the computer concept in the currently practiced understanding, that of Reduced Instruction Set Computer (RISC) type computing systems. Only for those interested and familiar with System Verilog HDL, at the end of this text there is an appendix that allows the simulation of the system described in this lesson.








Let us recall that Alan Turing, defining the machine that bears his name, divided the tape into two parts when he introduced the concept of the Universal Turing Machine (UTM). One part for describing the automaton of the Turing machine $M(T)$ and another for the associated tape, $T$. In other words, the Turing Machine automaton can be viewed as operating with two memories, one for programs and one for data. At the same time, mathematicians also developed the concept of multi-tape TM, which involves two or more tapes with the same number of read heads, which allows accessing multiple symbols stored in different memories in the same execution cycle.

{\bf A little history}: on 29 March 1944 John von Neumann, who was working on the Manhattan Project, run one of the first programs  on the {\bf Harvard Mark I} computer. The architect of this computer, Howard Aiken, decided for his computer an organization with two memories, one for programs and another for data. Thus, the emergence of the {\bf {\em Harvard abstract model}} for computers inspired, like the non-Neumann abstract model, from the mathematical concept of the Turing Machine is known in history. 

\section{Interpreting vs. Executing}

Let us to go back to the UTM concept. At this theoretical level, the processing is done as a "collaboration" between the "program" from the tape and the finite automaton with its finite number of states. Each change on the "data" part of the UTM's tape is performed in a number of steps involving (1) readings form the "program" part of tape and (2) state transitions of the UTM's finite automaton. The complexity of the computational process is given by
\begin{enumerate}
    \item the size of "program", a flexible structure because it can be redesigned as many time as it is needed
    \item the size of the description of the UTM's finite automaton, a fixed structure because because it is frozen in a physical implementation.
\end{enumerate}
From the designing point of view, we care about the size of the automaton, more precisely, about the complexity of the associated combinational circuit. Any hardware complexity deserves to be reduced due to the effort involved in {\em specification, design, verification, validation} and {\em testing}. Consequently, minimizing the complexity of the control provided, at the origin of the UTM, by the finite automaton is welcome. Reducing the number of states of the finite automaton does not entail a reduction in complexity. This was proven by Claude Shannon's approach \cite{Shannon56}, which reduced the number of states to only 2, but with the consequence of increasing the size of the binary circuit of the loop that calculates the next state. Only the definition of a finite automaton with zero states \cite{Stefan06}, therefore the elimination of the finite automaton, justifies, only belatedly, the elimination of the complexity introduced by the control.

Consequently, we distinguish two ways of acting on an instruction read from program memory:
\begin{itemize}
    \item the instruction is {\bf {\em interpreted}} by executing a sequence of micro-instructions generated by a control automaton (see \ref{contrAut} in this book)
    \item the instruction is {\bf {\em executed}} in a single clock cycle using the instruction fields directly applied to the processor components. 
\end{itemize}
It is obvious that, for an executive structure, we consider the suggestion offered by a UTM with two tapes and a processing system originating in a UTM with zero internal states.


\section{Harvard Computer Version}

The Harvard abstract model assumes a fifth loop. The processor together with the program memory forms a 4-OS, and by closing another loop through the data memory, a 5-OS is obtained (see Figure \ref{fig45}).

\placedrawing[H]{figures/13_01_harvard_model.lp}{The Harvard abstract model  of computer.}{fig45}

We will continue to focus, for obvious reasons, on RISC-type processors (see \ref{toyrisc} in this book) that allow the implementation of the Harvard abstract model.

\placedrawing[H]{figures/13_02_computer_organization.lp}{The organization of a Harvard version of a computing system.}{compSys}

The main difference compared with the von Neumann abstract version is due to the dual connection of the processor to the memory system. Tho independent connections, one for data and another for programs, allow the processor to access in the same cycle an instruction and, if needed, a data word to be processed in the RALU subsystem. In this way the memory hierarchy deepens with severe consequences for memory access speed. A miss in the L1 cache is resolved much faster because a much faster memory is used than the main one. The same effect is produced by a third level of cache if used.

\subsection{The Reduction Instruction Set Computer (RISC) Approach}

When computer scientists become armed with huge amounts of performance data, they were in the position to demonstrate that the simple design of a RISC system was able to easily outperform even the most powerful classic CPU designs, while at the same time producing machine code that was only marginally larger than the heavily optimized CISC (Complex Instruction Set Computer) instructions.

A reduction instruction set maintains only the simplest and  the most frequently used instructions from a large instruction set, provided that all rejected instructions can be implemented with the sequences of the remaining ones.

\subsection{Segregation Between the Simple Hardware and the Complex Program}

The Harvard version of the computer allows for a clear segregation of the simple hardware structure from the complex program structure. Hardware is build mainly from simple circuits. See the hardware components of the toyRISC (see \ref{toyrisc} in this book) processor where only few gates in the decoder section (see \ref{interrupt}) can be considered complex circuits, i.e., with the size in the same range with the description size.


\section{Cache Memories}

There are three  levels of cache memory in the currently designed processors: Level 1 (L1), Level 2 (L2), and Level 3 (L3).

A cache memory is a small, volatile type of computer memory that provides high-speed  access to data and programs stored in the virtual memory space.  

The first level of cache, L1, is the smallest but fastest cache memory. It's connected directly with the processor on the same silicon die, and typically ranges from 8KB to 128KB\footnote{This quantitative information evolves very rapidly over time due to technological advances, which is why at the time of publication of this volume, it is certainly outdated.}.  L1  provides the fastest speed because it operates at the same speed as the processor. In a Harvard version of a computer L1 is divided into two parts: one is data cache, L1d, and another is instructions cache, L1i.

L2 is larger than L1 and slightly slower, but still faster than main memory. It can be located also on the processor. It's typically used to store data that's not as frequently accessed as the data in L1. The size of L2 cache memory varies  ranging from 1MB to 8MB.

The third level, L3, is  larger and slower than L2, but it's still faster than main memory. L3 cache memory is designed to improve the speed of L1 and L2. It can be found on the motherboard or, increasingly, inside the processor. Its size  ranges from 8MB to 64MB or more.

In summary, the different levels of cache memory serve to bridge the gap between the high speed of processor  and the main memory's lower speed. They store, according to the {\em locality principle} (see \ref{locprinc} in this book), frequently used data and instructions, which helps to speed up data access and overall system performance. The levels, according to the {\em memory hierarchy} (see \ref{memhier} in this book), differ in size, speed, and location, with each level being larger, slower, and generally further from the CPU than the previous one.

\section{A Little History: Abstract Models for Computers}

The following formulations have become established for abstract computer models: von Neumann architecture and Harvard architecture. First, let's see what the established descriptions are.

\paragraph{Harvard Architecture} was implicitly introduced in the design of the Harvard Mark I, a relay-based computer developed at Harvard University in 1944. The Harvard architecture was not formally named as such at the time but referred to a system where separate memory units were used for program instructions and data. This concept was largely attributed to the work of Howard Aiken, who led the development of the Harvard Mark I. The architecture became formally recognized as a distinct model in computing later, particularly in the 1970s-1980s, especially in contexts involving RISC processors.

\paragraph{Von Neumann Architecture} gained common usage in the 1960s-1970s and was introduced by John von Neumann in the 1945 report titled "First Draft of a Report on the EDVAC" \cite{vonNeumann1945} \cite{vonNeumann45}. This document outlined the fundamental principles of a stored-program computer, where program instructions and data share the same memory. The report was written as part of a collaborative effort during the development of the EDVAC (Electronic Discrete Variable Automatic Computer) at the Moore School of Electrical Engineering, University of Pennsylvania. Although von Neumann's name is most associated with the architecture, the theoretical contributions of Alan Turing, and the contributions of J. Presper Eckert, John Mauchly, and others on the the ENIAC and EDVAC projects were also significant.

\bigskip

The two models cannot be considered architectures in the sense in which the concept of architecture is commonly used in computer science and technology. We believe that a correct name, instead of architecture, is {\bf {\em abstract model}}. An abstract model establishes the connection between a mathematical model and a concrete realization, a physical structure. So: the {\bf Harvard abstract model} and the {\bf von Neumann abstract model}.

\subsection{Facts about Harvard Mark I Computer}

The Harvard Mark I, or IBM Automatic Sequence Controlled Calculator (ASCC), was a general-purpose electromechanical computers whose original concept was presented to IBM by Howard Aiken in November 1937. He developed the concept during his master's and doctoral studies at Harvard when he was faced with solving differential equations that could only be solved numerically. The project was approved to be founded by the company chairman Thomas Watson Sr. in February 1939. The ASCC was developed and built by IBM at their Endicott plant, in the state of New York, and shipped to Harvard in February 1944.

We can conclude that it is a computer made by IBM based on the project of Howard Aiken, a doctoral student at Harvard University. {\em Aiken -- IBM -- Harvard} is the chain linked to the Mark I computer that draws our attention in the attempt to use a name for the abstract model in question. We cannot help but notice that "Harvard" appears at the end of the aforementioned series.


\subsection{Facts about "First Draft"}

At the Moore School of Electrical Engineering at the University of Pennsylvania, Herman Goldstine, John Mauchly, J. Presper Eckert and Arthur Burks began to study the development of the EDVAC, a computing machine thought as the follower of the ENIAC. A serendipitous encounter of Goldstine with von Neumann led to the "First Draft" memo hand written by von Neumenn and sent by mail to Goldstine, who typed in 24 copies and distributed it with John von Neumann as the sole author.  

It should be noted that the material was not considered by von Neumann to be finalized. The title and author were given by Goldstine. The text does not contain any bibliographical references, which could have been the author's unfinished intention. Let's credit the author with good intentions, as these references were perhaps intended to be inserted into the text. So, {\em we cannot accuse the author} of omitting certain contributions, because no reference was actually made to the origin of the ideas contained in the report. But, at the same time, we cannot avoid mentioning the contributions of von Neumann's direct collaborators, to which must be added the theoretical contribution of Alan Turing, whom von Neumann knew and appreciated because he proposed him to be an assistant in one of his courses.

\subsection{Renamings}

We believe that the current use of the terms "Harvard architecture" and "von Neuman architecture" is inappropriate for several reasons.

First, the concept of architecture (see \ref{archOrg} in this book) has a definition, imposed in the 1960s \cite{Amdahl64}, with a well-defined meaning related to the functional interface between hardware and software. For this reason, it is much more appropriate to use the term "abstract model" instead of "architecture".

Second, the association of the abstract model with the name of a famous university or a renowned mathematician is not appropriate according to what has been described above. It can always be proven that other names are more appropriate.

We believe that we will be able to use more appropriate names if we also take into consideration aspects related to the internal structure of the processors involved in structures with a single external memory as opposed to those with two external memories. The distinction made in \ref{archOrg} between {\em interpreting processor} and {\em executing processor} is significant for the difference between the two types of abstract models. The interpreting processor (CISC) is controlled by a complex automaton that allows and requires the expansion of each instruction read from memory into a {\em micro-program}, while the executive processor (RISC) can use a single execution cycle for each instruction because it can simultaneously access two memories, the program and the data.

From the perspective of the above, we allow ourselves to propose the substitution of the phrase "Harvard architecture" with that of "{\em executive abstract model}", and the phrase "von Neumann architecture" with that of "{\em interpretive abstract model}". We thus obtain names that reflect more significant structural aspects than those related only to the organization of memory external to the processor.






\section{{\bf {\em ToyRISC Processor}}}\label{toyrisc}

The executing processor is simpler than an interpreting processor.
The complexity of computation moves almost completely from the physical structure of the processor into the programs executed by the processor, because this kind of processor (called Reduced Instruction Set Computer, RISC)  has an organization containing mainly simple, recursively defined circuits.

\subsection{Organization}

The Harvard abstract model of a RISC executing machine  determines the internal structure of the
processor to have mechanisms allowing in each clock cycle cu
address both, the program memory and the data memory. Thus, the
RALU-type functional automaton, directly interfaced with the data
memory, is loop-connected with a control automaton designed to
fetch in each clock cycle a new instruction from the program
memory. The control automaton does not ``know" the function to be
performed, it ``knows"  only how to ``fetch the function" from an external storage
support, the program memory.

\placedrawing[H]{figures/05_06_toy_risc.lp}{The organization of the {\bf {\em toyRISC}}-based computer as a 5-OS. (To keep the image simple, the representation of command and control signals was avoided.)}{fig43}

{\bf Important note}: the organization presented in the following is designed only from the circuit point of view with emphasis on the {\em competence} of the system. In order to maximize the performance the design must apply specific techniques to increase the frequency of the clock cycle. This concern goes beyond the aims of this lecture. We will partially address these issues in the next lesson. Books on the organization and architecture of computing systems deal extensively with issues of maximizing performance. Books on the architecture and organization of computers written by John Hennessy and David Patterson are recommended  \cite{Hennessy19, Patterson19}.

In Figure \ref{fig43}, the organization of the processor {\bf {\em toyRISC}} is represented. It contains two simple loop-connected automata, PC and RALU, serially connected with a complex and small section, DCD. 


\subsubsection{PC: a control automaton}

The Control section is a simple functional automaton whose
state, stored in the register called Program Counter, ({\tt pc}), is used
to compute in each clock cycle the address in Program Memory from where the next
instruction is fetched. There are two modes to compute the next
address: (1) incrementing, with 1 or a signed number, the current
address, or (2) independently from the
current value of Program Counter, using a value fetched from an internal
register or a value generated by the currently executed
instruction.  More,
the current {\tt pc+1} can be stored in an internal register when
the control of the program {\em calls} a  function and a return is
needed. For all the previously described behaviors the
combinational circuit {\bf nextPC} (described in {\tt nextPc.v} module of the design presented in Appendix \ref{toyRiscDesign}) is designed to close the loop over the Program Counter register.


\subsubsection{RALU: a functional automaton}

The RALU section is also a simple functional automaton whose state is structured as an array of variables and is stored in a Register File (a small memory with two output ports and one input port). In each clock cycle two values are fetched from this memory and are submitted to be operated in the ALU unit. The result is write back in a location selected by a destination address. The
section accepts data coming form the data memory,
from the currently executed instruction, or from the {\bf PC}
automaton, thus closing the 3rd loop.

\subsubsection{DCD: interrupt automaton \& decoder}\label{interrupt}

The DCD section contains an automaton, the interrupt automaton, and a combinational decoder.

In  computers, an interrupt  is a signal for the processor to interrupt currently executing code. If the interrupt is accepted, the processor will suspend its current activities, save its state, and execute the function requested by the interrupt.  This interruption is  temporary, allowing the software to resume normal activities after the program associated to the interrupt finishes.

There are two types of interrupts:
\begin{itemize}
\item Hardware interrupts
when the interrupt signal generated from external devices (for example, in a keyboard if we press a key to do some action this pressing of the keyboard generates a signal that is given to the processor to do action). They are classified into two types:
\begin{itemize}
    \item Maskable Interrupt: interrupts that can be delayed when a highest priority interrupt has occurred to the processor.

    \item Non Maskable Interrupt: that cannot be delayed and immediately be serviced by the processor.
\end{itemize}
\item Software interrupts generated from internal devices  divided into two types:
\begin{itemize}
    \item Normal Interrupts: are caused by the software instructions are called software instructions.
    \item Exception: an unplanned interruption while executing a program (for example, while executing a program if we got a value that is divided by zero).
\end{itemize}
\end{itemize}

In our very simple example we illustrated a maskable hardware interrupt. The circuit {\tt intSect} consists of a finite automaton with two states, {\tt enableInterrupt} and {\tt disableInterrupt}, controlled with the instructions {\tt ei} and {\tt di}. When the system is reset, the machine switches to the {\tt disableInterrupt} state. If the automaton is in the {\tt enableInterrupt} state and the {\tt interrupt} signal is activated, then {\tt inta} (interrupt acknowledge) is generated, the {\tt pc+1} value is saved in {\tt regFile}, to be able to return the program to the point where it was interrupted, and  the instruction from {\tt intAddr} where the program associated with the interruption treatment is located is read. This program ends with an unconditional jump to the address saved in {\tt regFile}.

\bigskip


Both, the {\bf PC} automaton and the {\bf RALU} automaton are
simple, recursively defined automata. The computational complexity
is completely moved in the code stored inside the program memory.

\subsection{Instruction Set Architecture}

The storage resources on which the ISA definition is based are the following:
\begin{verbatim}
 reg    [31:0]  programMemory[0:65535]  ;
 reg    [31:0]  dataMemory[0:1023]      ;
 reg    [15:0]  pc                      ; // program counter
 reg    [31:0]  registers[0:31]         ; // register file
 reg            intEnable               ; // interrupt enable FF
 \end{verbatim}

 The structure of the arithmetic \& logic instructions have two forms:
 \begin{itemize}
    \item {\tt destinationRegiste <= leftOperamd OP rightOperand}
    \item {\tt destinationRegiste <= leftOperamd OP immediateValue}
 \end{itemize}
\hspace{-5mm}with the  following two formats:
\begin{verbatim}
        instruction[31:0] = {opCode[5:0],       // operation code
                             dest[4:0],         // destination register
                             left[4:0],         // left operand
                             right[4:0],        // right operand
                             noUse[10:0]}
                            |
                            {opCode[5:0],
                             dest[4:0],
                             left[4:0],
                             immediate[15:0]}   // value
\end{verbatim}

Instruction Set Architecture, ISA, of the {\bf {\em toyRISC}} processor is described in
the Listing \ref{lst:toyRISCdefines}. ISA, in short the architecture of a computing system includes at least the following types of instructions:
\begin{enumerate}
    \item control instructions
    \item arithmetic and logic instructions
    \item memory access instructions
    \item interrupt control instructions
\end{enumerate}

\includeverilogcode[caption={toyRISC ISA.  File: DEFINES.vh  (SystemVerilog)}, 
                    label={lst:toyRISCdefines}, 
                    firstline=1, 
                    firstnumber=1, 
                    lastline=40]
                    {\verilogSourceDir/A11_DEFINES.vh}

The first field of the instruction, {\tt opCode}, contains the information used to determine what is the the action performed by  the current instruction.
Each instruction is executed in one clock cycle. In Appendix \ref{toyRiscDesign} the entire design is listed. This design is meant to illustrate mainly the competence of the circuits without aiming to reach some performances in execution. In order to maximize the performance, it is necessary to apply special techniques (usually in the category of pipelines) related to the field of quantitative optimization of the organization of the computer system. See more on the quantitative approach in \citep{Hennessy19}.

\subsection{Assembly Code}

In order to generate the code executable by the toyRISC processor, a code generator  translates the mnemonics into binary code. In Table \ref{mnem}, the mnemonics used to write assembly programs are listed. In the second column the action performed specified and in the last column the binary codes are listed. The binary code is organized in 5 or 4 fields.  The first field represents the operation code, the second the destination address of the result  in the register file ({\tt ddddd} is the binary code of the actual value when it matters), the third field represents the address for the left operand ({\tt lllll} is the binary code of the actual value when it matters). The least significant 16 bits are organized in 2 fields or in one field. In the first case 5 bits ({\tt rrrrr} is the binary code of the actual value when it matters) represents the address of the right operand and the rest 11 bits have no meaning, while in the second case all the 16 bits represents a value ({\tt vvvvvvvvvvvvvvvv} is the binary code of the actual value when it matters).

\begin{table}[H]
\footnotesize
\centering
\begin{longtable}{|l|l|l|}
\caption{Assembly language mnemonics, their meaning and binary form.}
\label{tab:riscISA} \\
\hline
\textbf{Mnemonics} & Description & Binary form \\
\hline
\endfirsthead
\hline
\textbf{Mnemonics} & Description & Binary form \\
\hline
\endhead
\hline
\endfoot
\hline
\endlastfoot
      NOP                       & rf[0] <= rf[0] + 0                            & 110010\_00000\_00000\_00000\_00000000000  \\ \hline
      RJMP(label)               & pc <= pc + value                              & 000001\_00000\_00000\_vvvvvvvvvvvvvvv     \\ \hline
      BRZ(left,label)         & pc <= (left = 0) ? pc+immediate : pc+1          & 000010\_00000\_lllll\_vvvvvvvvvvvvvvv     \\ \hline
      BRNZ(left,label)        & pc <= (left = 0) ?  pc+1 : pc+immediate         & 000011\_00000\_lllll\_vvvvvvvvvvvvvvv     \\ \hline
      RET                       & pc <= rf[left]                                & 000101\_00000\_lllll\_00000\_00000000000  \\ \hline
      HALT                      & pc <= pc                                      & 000110\_00000\_00000\_00000\_00000000000  \\ \hline
      EINT                      & intEnable <= 1 (enable interrupt)             & 001000\_00000\_00000\_00000\_00000000000  \\ \hline
      DINT                      & intEnable <= 0 (disable interrupt)            & 001001\_00000\_00000\_00000\_00000000000  \\ \hline
      ADD(dest,left,right)    & rf[dest] <= rf[left] + rf[right]                & 110000\_ddddd\_lllll\_rrrrr\_00000000000  \\ \hline
      SUB(dest,left,right)    & rf[dest] <= rf[left] - rf[right]                & 110001\_ddddd\_lllll\_rrrrr\_00000000000  \\ \hline
      ADDV(dest,left,value)    & rf[dest] <= rf[left] + value                   & 110010\_ddddd\_lllll\_vvvvvvvvvvvvvvv     \\ \hline
      MULT(dest,left,right)   & rf[dest] <= rf[left] * rf[right]                & 110011\_ddddd\_lllll\_rrrrr\_00000000000  \\ \hline
      MULTV(dest,left,value)   & rf[dest] <= rf[left] * rf[right]               & 110100\_ddddd\_lllll\_vvvvvvvvvvvvvvv     \\ \hline
      ADDC(dest,left,right)  & rf[dest] <= (rf[left] + rf[right])[32]           & 110101\_ddddd\_lllll\_rrrrr\_00000000000  \\ \hline
      SUBC(dest,left,right)  & rf[dest] <= (rf[left] - rf[right])[32]           & 110110\_ddddd\_lllll\_rrrrr\_00000000000  \\ \hline
      ADDVC(dest,left,value)  & rf[dest] <= (rf[left] + value)[32]              & 110111\_ddddd\_lllll\_vvvvvvvvvvvvvvv     \\ \hline
      LSH(dest,left)        & rf[dest] <= \{0, rf[left][31:1]\}                 & 111000\_ddddd\_lllll\_00000\_00000000000  \\ \hline
      ASH(dest,left)        & rf[dest] <= \{rf[left][31],rf[left][31:1]\}       & 111001\_ddddd\_lllll\_00000\_00000000000  \\ \hline
      MOVE(dest,left)       & rf[dest] <= rf[left]                              & 111010\_ddddd\_lllll\_00000\_00000000000  \\ \hline
      SWAP(dest,left)       & rf[dest] <= {rf[left][15:0],rf[left][31:16]}      & 111011\_ddddd\_lllll\_00000\_00000000000  \\ \hline
      NOT(dest,left)        & rf[dest] <= ~rf[left]                             & 111100\_ddddd\_lllll\_00000\_00000000000  \\ \hline
      AND(dest,left,right)    & rf[dest] <= rf[left \& rf[right]                & 111101\_ddddd\_lllll\_rrrrr\_00000000000  \\ \hline
      OR(dest,left,right)     & rf[dest] <= rf[left] | rf[right]                & 111110\_ddddd\_lllll\_rrrrr\_00000000000  \\ \hline
      XOR(dest,left,right)    & rf[dest] <= rf[left] $\oplus$ rf[right]         & 111111\_ddddd\_lllll\_rrrrr\_00000000000  \\ \hline
      READ(left)            & read from dataMemory[left]                        & 100000\_00000\_lllll\_00000\_00000000000  \\ \hline
      LOAD(dest)            & rf[dest] <= dataOut                               & 100111\_ddddd\_00000\_00000\_00000000000  \\ \hline
      STORE(left,right)     & dataMemory[left] <= rf[right]                     & 101000\_00000\_lllll\_rrrrr\_00000000000  \\ \hline
      VAL(dest,val)         & rf[dest] <= \{{11{val[20]}}, value\}              & 010111\_ddddd\_00000\_vvvvvvvvvvvvvvv     \\ \hline
\end{longtable}
\end{table}

\subsubsection{Toy Assembler}

The binary codes in the \ref{tab:riscISA} table are generated, for reasons of maintaining a simple design and simulation environment, using a program also written in System Verilog. This program, {\tt RISCcodeGenerator.sv}, receives a sequence of instructions in the file {\tt program.sv} and generates the executable binary code in the program memory {\tt progMemory}. This generator is listed  in the \ref{codeGen} section (see Listing \ref{lst:RISCcodeGenerator}). The line:
\begin{verbatim}
        `include "program.sv"
\end{verbatim}
Is included twice on lines 370 and 372.
In the first pass, the absolute positions of the labels are identified through the {\tt LB} task in the {\tt labelTab} table, and in the second pass through the {\tt ULB} task, the relative jump is calculated, which is correctly completed in the code binary. The two tasks are necessary because some relative jumps are at advanced locations compared to the position of the jump instruction.

\subsubsection{Simulator}

The simulator used to validate the processor design is described in detail in the \ref{riscsim} section, where the processor was instantiated and the data, {\tt dataMemory}, and program, {\tt ProgMemory} memories were added. The interrupt signal {\tt intIn} has been activated, which will be taken into account immediately by the instruction {\tt DI} will allow it.


\subsubsection{Assembly Programs}

\begin{exemplu}\label{exADD}
Let by a following simple program:
{\em 
\includeverilogcode[caption={toyRISC program: initializes the first 5 registers and adds them in register 0.  File: A11\_RISCprogram1.sv  (SystemVerilog)}, 
                    label={lst:A11_RISCprogram1.sv}, 
                    firstline=1, 
                    firstnumber=1, 
                    lastline=30]
                    {\verilogSourceDir/A11_RISCprogram1.sv}
}

The code generator loads in the program memory the following binary form of the program:
{\small
\begin{verbatim}
progMemory[0] 	 = 01011100000000000000000000000001
progMemory[1] 	 = 01011100001000000000000000000010
progMemory[2] 	 = 01011100010000000000000000000011
progMemory[3] 	 = 01011100011000000000000000000100
progMemory[4] 	 = 01011100100000000000000000000101
progMemory[5] 	 = 11000000000000000000100000000000
progMemory[6] 	 = 11000000000000000001000000000000
progMemory[7] 	 = 11000000000000000001100000000000
progMemory[8] 	 = 11000000000000000010000000000000
\end{verbatim}
}
The clock period is equal with 2 time units. The program starts at the time unit {\tt t=5} after the 4 time units action of the {\tt reset} signal.
The value of the program counter and of the first 2 registers in the register file evolve as follows:
{\small
\begin{verbatim}
pc= 0 RF=[1, x, x, x]  ei=0 inta=0
pc= 1 RF=[1, x, x, x]  ei=0 inta=0
pc= 2 RF=[1, 2, x, x]  ei=0 inta=0
pc= 3 RF=[1, 2, 3, x]  ei=0 inta=0
pc= 4 RF=[1, 2, 3, 4]  ei=0 inta=0
pc= 5 RF=[1, 2, 3, 4]  ei=0 inta=0
pc= 6 RF=[3, 2, 3, 4]  ei=0 inta=0
pc= 7 RF=[6, 2, 3, 4]  ei=0 inta=0
pc= 8 RF=[10, 2, 3, 4] ei=0 inta=0
pc= 9 RF=[15, 2, 3, 4] ei=0 inta=0
\end{verbatim}
}

$\diamond$
\end{exemplu}

\begin{exemplu}
Let us take an example to show how the interrupt acts. The interrupt is applied from the beginning of the test, but it is enabled only after the instruction {\tt EI} runs.
{\em 
\includeverilogcode[caption={toyRISC program: tests the interrupt action.  File: A11\_RISCprogram2.sv  (SystemVerilog)}, 
                    label={lst:A11_RISCprogram2.sv}, 
                    firstline=1, 
                    firstnumber=1, 
                    lastline=30]
                    {\verilogSourceDir/A11_RISCprogram2.sv}
}

The toy assembler generates in {\tt progMemory} the following executable code:

{\small
\begin{verbatim}
progMemory[0]    = 01011111111000000000000000001010
progMemory[1]    = 01011100010000000000000000010111
progMemory[2]    = 01011100000000000000000000001101
progMemory[3]    = 00100000000000000000000000000000
progMemory[4]    = 11001000000000000000000000000010
progMemory[5]    = 11001000000000000000000000000000
progMemory[6]    = 11001000000000000000000000000100
progMemory[7]    = 00011000000000000000000000000000
progMemory[8]    = 11001000000000000000000000000000
progMemory[9]    = 11001000000000000000000000000000
progMemory[10] 	 = 00100100000000000000000000000000
progMemory[11] 	 = 01011100011000000000000000101100
progMemory[12] 	 = 00010100000111100000000000000000
progMemory[13] 	 = xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx
\end{verbatim}
}
The simulation provides:
{\small
\begin{verbatim}
pc=  0  RF=[x, x, x, x]     ei=0 inta=0
pc=  1  RF=[x, x, x, x]     ei=0 inta=0
pc=  2  RF=[x, x, 23, x]    ei=0 inta=0
pc=  3  RF=[13, x, 23, x]   ei=0 inta=0
pc=  4  RF=[13, x, 23, x]   ei=1 inta=1
pc= 10  RF=[13, x, 23, x]   ei=0 inta=0
pc= 11  RF=[13, x, 23, x]   ei=0 inta=0
pc= 12  RF=[13, x, 23, 44]  ei=0 inta=0
pc=  4  RF=[13, x, 23, 44]  ei=0 inta=0
pc=  5  RF=[15, x, 23, 44]  ei=0 inta=0
pc=  6  RF=[15, x, 23, 44]  ei=0 inta=0
pc=  7  RF=[19, x, 23, 44]  ei=0 inta=0
\end{verbatim}
}

$\diamond$
\end{exemplu}

\begin{exemplu}
Let be a wrong program which runs forever because of an unconditional jump.
{\em 
\includeverilogcode[caption={toyRISC program: an unending running program.  File: A11\_RISCprogram3.sv  (SystemVerilog)}, 
                    label={lst:A11_RISCprogram3.sv}, 
                    firstline=1, 
                    firstnumber=1, 
                    lastline=30]
                    {\verilogSourceDir/A11_RISCprogram3.sv}
}

The simulation provides:
{\small
\begin{verbatim}
pc= 0  RF=[3, x, x, x]          ei=0 inta=0
pc= 1  RF=[3, x, x, x]          ei=0 inta=0
pc= 2  RF=[2, x, x, x]          ei=0 inta=0
pc= 3  RF=[2, x, x, x]          ei=0 inta=0
pc= 1  RF=[2, x, x, x]          ei=0 inta=0
pc= 2  RF=[1, x, x, x]          ei=0 inta=0
pc= 3  RF=[1, x, x, x]          ei=0 inta=0
pc= 1  RF=[1, x, x, x]          ei=0 inta=0
pc= 2  RF=[0, x, x, x]          ei=0 inta=0
pc= 3  RF=[0, x, x, x]          ei=0 inta=0
pc= 1  RF=[0, x, x, x]          ei=0 inta=0
pc= 2  RF=[4294967295, x, x, x] ei=0 inta=0
pc= 3  RF=[4294967295, x, x, x] ei=0 inta=0
pc= 1  RF=[4294967295, x, x, x] ei=0 inta=0
pc= 2  RF=[4294967294, x, x, x] ei=0 inta=0
pc= 3  RF=[4294967294, x, x, x] ei=0 inta=0
pc= 1  RF=[4294967294, x, x, x] ei=0 inta=0
pc= 2  RF=[4294967293, x, x, x] ei=0 inta=0
pc= 3  RF=[4294967293, x, x, x] ei=0 inta=0
pc= 1  RF=[4294967293, x, x, x] ei=0 inta=0
\end{verbatim}
}

$\diamond$
\end{exemplu}


\begin{exemplu}
Let be a  program which works with {\tt dataMemory}.
{\em 
\includeverilogcode[caption={toyRISC program: tests the access to the data memory.  File: A11\_RISCprogram4.sv  (SystemVerilog)}, 
                    label={lst:A11_RISCprogram4.sv}, 
                    firstline=1, 
                    firstnumber=1, 
                    lastline=30]
                    {\verilogSourceDir/A11_RISCprogram4.sv}
}

The simulation provides:
{\small
\begin{verbatim}
pc= 0  RF=[1, x, x, x]    ei=0 inta=0
pc= 1  RF=[1, x, x, x]    ei=0 inta=0
pc= 2  RF=[1, 55, x, x]   ei=0 inta=0
pc= 3  RF=[1, 55, x, x]   ei=0 inta=0
pc= 4  RF=[1, 55, x, x]   ei=0 inta=0
pc= 5  RF=[1, 55, 55, x]  ei=0 inta=0
\end{verbatim}
}

$\diamond$
\end{exemplu}




\subsection{Time performance}

The longest combinational path in a system using our {\bf {\em
toyRISC}}, which imposes the minimum clock period, is:

$$T_{clock} = t_{clock\_to\_instruction} + t_{leftAddr\_to\_leftOp}
+ t_{throughALU} + t_{throughMUX} + t_{fileRegSU}$$

Because the system is not buffered the clock frequency depends also by the
time behavior of the system directly connected with {\bf {\em
toyRISC}}. In this case $t_{clock\_to\_instruction}$ -- the access
time of the program memory, related to the active edge of the
clock -- is an extra-system parameter limiting the speed of our
design. The internal propagation time to be considered are: the
read time from the file register ($t_{leftAddr\_to\_leftOp}$ or
$t_{rightAddr\_to\_rightOp}$), the maximum propagation time
through ALU (dominated by the time for an 32-bit arithmetic
operation), the propagation time through a 4-way 32-bit
multiplexer, and the {\em set-up time} on the file register's data
inputs. The way from the output of the file register through {\bf
Next PC} circuit is ``shorter" because it contains a 16-bit adder,
comparing with the 32-bit one of the ALU.

The increase in speed performance will be illustrated in the next lesson by using pipeline registers.

\section{How is Designed an Instruction Set Architecture}

As we already said, we use the term ISA to refer to the actual
programmer-visible instruction set. The ISA serves as the boundary
between the software engineer  and hardware engineer. We will list and comment further on the main characteristics of an ISA 
% FIXME: Not in bib \cite{Patterson4}:
\begin{itemize}
    \item Class of ISA
        \begin{itemize}
            \item register-memory ISAs such as x86, with complex memory access instructions
            \item load-store ISAs such as ARM, RISC V, with only simple load and store insructions for memory access
        \end{itemize}
    \item Memory addressing is byte addressing with two versions:
        \begin{itemize}
            \item byte aligned (ex.: ARM) if the object accessed at address A has $s$ bytes, then the  $A mod s = 0$
            \item with no alignment (ex.: x86 and RISC V) which produces slower access
        \end{itemize}
    \item Addressing modes
        \begin{itemize}
            \item few simple: register, immediate, displacement, ... (RISC V, ARM)
            \item many complex: the previous and more (x86)
        \end{itemize}
    \item Types and sizes of operands
        \begin{itemize}
            \item integer: 8 bit (ASCII), 16 bit (Unicode character), 32-bit (word), 64-bit (double word)
            \item floats: standard IEEE 754 32-bit floating point and 64-bit floating point
        \end{itemize}
    \item Operations:
        \begin{itemize}
            \item data transfer
            \item arithmetic
                \begin{itemize}
                    \item integer: add, sub, mult, div, rem, rightShift, ...
                    \item floats: add, sub, mult, div, rem
                \end{itemize}
            \item logic: minimally AND and XOR
            \item control: jumps, conditional branches, calls, ret, ...
        \end{itemize}
    \item Control flow instructions:
        \begin{itemize}
            \item RISC V tests the value of a register for conditional branches
            \item x86 and ARM test flags in a state register
            \item x86 save the return address from subroutine on a stack memory organized in the main mamory
        \end{itemize}
    \item Encoding an ISA:
        \begin{itemize}
            \item variable length for x86
            \item fix length for RISC V and ARM
        \end{itemize}
\end{itemize}


\section{Problems}

\begin{problem}
Install on your computer the design of the toyRISC processor using modules listed in Section \ref{toyRiscDesign}, and run the simple program which loads in the first 4 registers the numbers 11, 22, 33, 44.

$\diamond$
\end{problem}

{\bf Solution}

The program is:
\includeverilogcode[caption={toyRISC program used to validate the simulator.  File: A11\_RISCprogram5.sv  (SystemVerilog)}, 
                    label={lst:A11_RISCprogram5.sv}, 
                    firstline=1, 
                    firstnumber=1, 
                    lastline=30]
                    {\verilogSourceDir/A11_RISCprogram5.sv}

The program generated by {\tt RISCcodeGenerator.sv} is:
{\small
\begin{verbatim}
progMemory[0]    = 01011100000000000000000000001011
progMemory[1]    = 01011100001000000000000000010110
progMemory[2]    = 01011100010000000000000000100001
progMemory[3]    = 01011100011000000000000000101100
progMemory[4]    = 00011000000000000000000000000000
progMemory[5]    = xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx
\end{verbatim}
}
The result of simulation is:
{\small
\begin{verbatim}
pc= 0  RF=[11, x, x, x]
pc= 1  RF=[11, x, x, x]
pc= 2  RF=[11, 22, x, x]
pc= 3  RF=[11, 22, 33, x]
pc= 4  RF=[11, 22, 33, 44]
\end{verbatim}
}

\begin{problem}
Add to the list of monitoring signals the last two register in the register file in order to verify the way the interrupt signal is managed. Write a program which contains the subroutine of 5 instruction at the location 20 in the program memory. Enable the interrupt action in the sixth step of the main program. Make also all the necessary adjustment in module {\tt testRISC.sv} to allow verifying the interrupt action.

$\diamond$
\end{problem}


\begin{problem}
The add-prefix function receives the sequence of numbers $[n_0, n_1, \ldots, n_{m-1}]$ and returns $$[n_0, n_0 + n_1, n_0 + n_1 + n_2, \ldots, \sum_0^{m-1} n_i]$$
Write the program that load in the first 16 registers a sequence of different numbers and compute the add-prefix sequence in the last 16 registers.

$\diamond$
\end{problem}


\begin{problem}\label{div}
In registers 0 and 1 are stored the positive integers A an B. Write in assembly the program that compute in register 2 the quotient $A / B$ and 3 the reminder.

{\bf Hint}: subtract as many times as possible B from A counting in 2 successes. In register 3 leave the rest.

$\diamond$
\end{problem}
What instruction or instructions should be added to the architecture to simplify the program in Problem \ref{div}?

\begin{problem}
Load in data memory the first 32 Fibbonacci numbers and then add them in register 3.


$\diamond$
\end{problem}

